---
title: "Data Cleaning"
author: "Jack A. Goldman"
date: "2025-12-26"
format: html
---


```{r}
# Load necessary libraries
library(dplyr)
library(tidyverse)
library(readr)
library(zoo)
library(here)

# load fire information
fire_info <- read_csv(here("data", "bsw_fire_information.csv")) 
```


# Purpose 

The purpose of this script is to clean and preprocess the raw dataset for further analysis. 

# VPD data cleaning for lagged analysis
```{r}
# Load the raw dataset
vpd_data <- read_csv(here("data", "bsw_daily_prefire_vpd.csv"))
# Inspect the data
head(vpd_data)
str(vpd_data)
```

Filter fire info to only include GIDs that are in vpd_data

```{r}
fire_info_filtered <- fire_info |> 
    filter(GID %in% unique(vpd_data$GID))

```

For each GID filter date so that we only have dates within 30 days before fire start date (fire starte date is the start_date column in fire_info) and add column lag that is 1-30 for days before fire start date


```{r}
vpd_30 <- vpd_data |> 
    left_join(fire_info |> select(GID, start_date), by = "GID") |> 
    mutate(
        date = as.Date(date),
        start_date = as.Date(start_date)
    ) |> 
    filter(date >= (start_date - 30) & date < start_date) |>  # keep 1–30 days before start
    mutate(lag = as.integer(start_date - date)) |>            # lag = 1..30
    select(-start_date)

```


VPD matrix for distributed lag model
Currently in long format, need to convert to wide format with each column as vpd for a specific lag and each row as a fire event (GID) but arrange columns by lag vpd_lag_1 to vpd_lag_30
```{r}
vpd_wide = vpd_30 %>%
    select(GID, lag, vpd) %>%
    distinct(GID, lag, .keep_all = TRUE) %>%
    arrange(lag) %>%
    pivot_wider(names_from = lag, values_from = vpd, names_prefix = "vpd_") %>%
    drop_na() |> 
    arrange(GID)

vpd_wide_gids = unique(vpd_wide$GID)
nrow(vpd_wide)
```

Save vpd_wide as vpd_30_wide.csv
```{r}
saveRDS(vpd_wide, here("data", "cleaned", "vpd_30_wide.rds"))
```

# vpd data cleaning for moving average over 30 days
```{r}

vpd_ma_30 = vpd_data |> 
    left_join(fire_info |> select(GID, start_date), by = "GID") |> 
    mutate(
        date = as.Date(date),
        start_date = as.Date(start_date)
    ) |> 
    arrange(GID, date) |> 
    group_by(GID) |> 
    filter(date >= (start_date - 30) & date < start_date) |> 
    mutate(vpd_ma_30 = zoo::rollapply(vpd, width = 30, FUN = function(x) mean(x, na.rm = TRUE), fill = NA, align = "right")) |> 
    summarize(
        vpd_ma_30 = { tmp <- na.omit(vpd_ma_30); if (length(tmp)) tmp[length(tmp)] else NA_real_ },
        .groups = "drop"
    ) |> 
    drop_na() 
```

Save vpd_ma_30 

```{r}
saveRDS(vpd_ma_30, here("data", "cleaned","vpd_ma_30.rds"))
```

 # RBR data cleaning
 Load RBR zonal stats data and select only GID, median, and 90th percentile

 ```{r}
rbr_data = read_csv(here("data", "batch_rbr_w_offset_zonal_stats.csv")) |> 
    select(GID, median, percentile_90)  |> 
    rename(pct_90 = percentile_90) 

 ```

 Save cleaned RBR data
 ```{r}
saveRDS(rbr_data, here("data", "cleaned/rbr_cleaned.rds"))
 ```

# Landcover cleaning


```{r}
landcover_data = read_csv(here("data", "fire_landcover_pct_v1.csv")) 

head(landcover_data)

landcover_cleaned = landcover_data |> 
    select(GID, coniferous, percent_upland, percent_wetland) |> 
    rename(pct_conf = coniferous, 
           pct_up = percent_upland, 
           pct_wet = percent_wetland)

saveRDS(landcover_cleaned, here("data", "cleaned", "landcover_cleaned.rds"))
```

# fire information cleaning

```{r}
fire_info_cleaned = fire_info |> 
        select(GID, YEAR, start_date, ecoregion, ecoregion_lvl, size_ha, lat, lon) |> 
        rename(fire_size_ha = size_ha, 
               year = YEAR, 
               start_date = start_date) |> 
        mutate(month = as.integer(format(as.Date(start_date), "%m")))

```

if fire burned before june 1 set it as spring fire season else summer fire season 
The reason for this is because aspen/deciduous leaf out changes the flammability of the landscape and has been shown to impact fire intensity/severity between spring/summer (Podur and Martell 2009; Parisien et al. 2023)

```{r}
fire_info_cleaned <- fire_info_cleaned |> 
    mutate(season = ifelse(month < 6, "spring", "summer"))
```

save the file

```{r}
saveRDS(fire_info_cleaned, here("data", "cleaned", "fire_info_cleaned.rds"))
```


# cmi data cleaning

Read in cmi data


```{r}
cmi_data = read_csv(here("data", "cmi_hogg_water_year_anomaly.csv"))

```


```{r}
str(cmi_data)
```

Join fire info by GID
```{r}
cmi_data <- cmi_data |> 
    left_join(fire_info |> select(GID, YEAR), by = "GID") 
```

For each fire, remove water year rows with match YEAR (fire year) so that there are only water years before the fire year
```{r}
cmi_cleaned <- cmi_data |> 
    filter(water_year < YEAR) |> 
    select(GID, water_year, YEAR, avg_cmi_anomaly) |> 
    rename(cmi_wy_anom = avg_cmi_anomaly)
```

If water year is YEAR - 1 then lag = 1, if water year is YEAR - 2 then lag = 2, etc.
```{r}
cmi_cleaned <- cmi_cleaned |> 
    mutate(lag = YEAR - water_year) |> 
    select(-YEAR) |>  
    filter(lag <= 5) |> # keep only lags 1-5
    select(-c(water_year)) |> 
    arrange(GID, lag)
```

Pivot wider so that each column is cmi lag 1 to lag 5
```{r}
cmi_wide <- cmi_cleaned |> 
    distinct(GID, lag, .keep_all = TRUE) %>%
    pivot_wider(names_from = lag, values_from = cmi_wy_anom, names_prefix = "cmi_wy_") |> 
    drop_na() |> 
    arrange(GID)
```

# calculate a moving average of cmi_wy_1 to cmi_wy_5 and add as a new column cmi_wy_ma_5 using zoo::rollapply using cmi_cleaned
```{r}
cmi_ma <- cmi_cleaned |> 
    group_by(GID) |> 
    arrange(lag) |> 
    summarize(
        cmi_wy_ma_5 = { tmp <- zoo::rollapply(cmi_wy_anom, width = 5, FUN = function(x) mean(x, na.rm = TRUE), fill = NA, align = "right"); if (length(na.omit(tmp))) na.omit(tmp)[length(na.omit(tmp))] else NA_real_ },
        .groups = "drop"
    ) |> 
    drop_na()
```

# save cmi_wide joined 
```{r}
saveRDS(cmi_wide, "data/cleaned/cmi_wy_cleaned.rds")
```


# save cmi_ma
```{r}
saveRDS(cmi_ma, here("data", "cleaned","cmi_wy_ma_5.rds"))
```


# CMI 36

```{r}
long_term_cmi <- read_csv(here("data", "cleaned","cmi_hogg_36_months_before_fire.csv"))
```