---
title: "Median DLNM Analysis"
author: "Jack A. Goldman"
date: "2025-12-26"
format: html
---

This use mgcv and dlnm packages to fit a distributed lag model for VPD and fire severity (RBR median)
We use a matrix of VPD values for each fire event (GID) with columns representing lags from 1 to 30 days before fire start date

# Objective
- Short-term increases in VPD (drying conditions) in the days leading up to fire ignition will lead to higher burn severity (RBR median)
- Long-term VPD trends (e.g., seasonal averages) may have less influence on burn severity compared to short-term fluctuations by decreasing fuel abundance

- lagged effects of VPD on burn severity will be non-linear, with immediate effects being stronger than delayed effects
- lagged effects of  CMI on burn severity will be non-linear, with immediate effects being stronger than delayed effects


The goal is to model the effects of short-term and long-term moisture conditions on median and extreme burn severity as having a non-linear, lagged effect and compare whether short-term or long-term moisture conditions better explain burn severity patterns.

# Questions

To achieve this goal, I'll test 4 hypotheses for short and long-term moisture conditions (VPD and CMI (maybe RH and FWI?)) on both median and extreme (95th percentile) burn severity using a series of GAMs with distributed lag non-linear models implemented using the `mgcv`

1.  What is the relative impact of moisture conditions on burn severity?
2.  Are short-term or long-term moisture conditions more important?
3.  Are relationships consistent between median and extreme?
4.  Where are there lagged effects?



```{r}
library(ggplot2)
library(tidyr)
library(ggplot2)
library(viridis)
library(cowplot)
library(patchwork)
library(readr)
library(mgcv)
library(dlnm)
library(here)
library(gratia)
```


# Read in data

```{r}
moisture_data <- readRDS(here ("data", "analysis_ready","moisture_lagged_data.rds")) 
```

```{r}
#are there nas in the data
sum(is.na(moisture_data))
moisture_data <- moisture_data |>  
  filter(fire_size_ha > 200) |> 
  filter(ecoregion_lvl %in%c("LN", "BTL", "LSJ"))
```

# Build Full Models



## vpd with distributed lag
```{r}

m_vpd_median <- gam(median ~ 
       te(vpd, L,  by = weights_btl, k = c(3,4), bs = c('tp', 'cr')) + # vpd with lag
       te(vpd, L,  by = weights_ln, k = c(3,4), bs = c('tp', 'cr')) + # vpd with lag
       te(vpd, L,  by = weights_lsj, k = c(3,4), bs = c('tp', 'cr')) + # vpd with lag
         pct_conf + fwi_90th  +
         mean_evi +  
        s(fyear,bs = "re") +
        s(ecoregion_lvl, bs = "re") ,
      data = moisture_data,
      family = scat(link = "identity"),
      select = TRUE, 
      method = "REML")

```

```{r}
m_vpd_ext <- gam(pct_95 ~ 
       te(vpd, L,  by = weights_btl, k = c(3,4), bs = c('tp', 'cr')) + # vpd with lag
       te(vpd, L,  by = weights_ln, k = c(3,4), bs = c('tp', 'cr')) + # vpd with lag
       te(vpd, L,  by = weights_lsj, k = c(3,4), bs = c('tp', 'cr')) + # vpd with lag
         pct_conf+ fwi_90th  +
         mean_evi +  
        s(fyear,bs = "re") +
        s(ecoregion_lvl, bs = "re") ,
      data = moisture_data,
      family = scat(link = "identity"),
      select = TRUE,
      method = "REML")
```

## cmi

## cmi with distributed lag
```{r}

m_cmi_median <- gam(median ~ 
       te(cmi_wy, LL,  by = weights_btl2, k = c(3,4), bs = c('tp', 'cr')) + # vpd with lag
       te(cmi_wy, LL,   by = weights_ln2, k = c(3,4), bs = c('tp', 'cr')) + # vpd with lag
       te(cmi_wy, LL,   by = weights_lsj2, k = c(3,4), bs = c('tp', 'cr')) + # vpd with lag # cmi with 5 year lag
        pct_conf + 
        fwi_90th + 
        mean_evi + 
        s(fyear,bs = "re") +
        s(ecoregion_lvl, bs = "re") ,
      data = moisture_data,
      family = scat(link = "identity"),
      select = TRUE,
      method = "REML")

```

```{r}
m_cmi_ext <- gam(pct_95 ~ 
       te(cmi_wy, LL,  by = weights_btl2, k = c(3,4), bs = c('tp', 'cr')) + # vpd with lag
       te(cmi_wy, LL,   by = weights_ln2, k = c(3,4), bs = c('tp', 'cr')) + # vpd with lag
       te(cmi_wy, LL,   by = weights_lsj2, k = c(5,5), bs = c('tp', 'cr')) + # vpd with lag # cmi with 5 year lag
        pct_conf + 
        fwi_90th + 
        mean_evi + 
        s(fyear,bs = "re") +
        s(ecoregion_lvl, bs = "re") ,
      data = moisture_data,
      family = scat(link = "identity"),
      select = TRUE,
      method = "REML")
```

## save models

```{r}

saveRDS(m_vpd_median, here("analysis", "models","m_vpd_median.rds"))
saveRDS(m_cmi_median, here("analysis", "models","m_cmi_median.rds"))
saveRDS(m_vpd_ext, here("analysis", "models","m_vpd_ext.rds"))
saveRDS(m_cmi_ext, here("analysis", "models","m_cmi_ext.rds"))
```

## assess model fits

vpd
```{r}
gam.check(m_vpd_median)
```

cmi
```{r}
gam.check(m_cmi_median)

```

# H1: What is the relative impact of moisture conditions on burn severity?

checked hierarchical partitioning using gam.hp

## median
```{r}
iv <- list(
  moisture = c("te(vpd,L,by=weights_btl,k=c(3,4),bs=c(\"tp\",\"cr\"))", "te(vpd,L,by=weights_ln,k=c(3,4),bs=c(\"tp\",\"cr\"))", "te(vpd,L,by=weights_lsj,k=c(3,4),bs=c(\"tp\",\"cr\"))"), 
  fire_weather = "fwi_90th",
  fuels = c("pct_conf", "mean_evi"))
vpd_median_hp <- gam.hp(m_vpd_median, iv)
vpd_median_hp
```

Moisture contributes most to the model, explained 57% of the deviance compared to 39% for fuels and 2.1% for fire weather

```{r}
iv <- list(
  moisture = c("te(cmi_wy,LL,by=weights_btl2,k=c(3,4),bs=c(\"tp\",\"cr\"))", "te(cmi_wy,LL,by=weights_ln2,k=c(3,4),bs=c(\"tp\",\"cr\"))", "te(cmi_wy,LL,by=weights_lsj2,k=c(3,4),bs=c(\"tp\",\"cr\"))"), 
  fire_weather = "fwi_90th",
  fuels = c("pct_conf", "mean_evi"))
cmi_median_hp <- gam.hp(m_cmi_median, iv)
cmi_median_hp
```

Moisture contributes most to the model, explained 62% of the deviance compared to 36% for fuels and 1.8% for fire weather


## Extreme

```{r}
iv <- list(
  moisture = c("te(vpd,L,by=weights_btl,k=c(3,4),bs=c(\"tp\",\"cr\"))", "te(vpd,L,by=weights_ln,k=c(3,4),bs=c(\"tp\",\"cr\"))", "te(vpd,L,by=weights_lsj,k=c(3,4),bs=c(\"tp\",\"cr\"))"), 
  fire_weather = "fwi_90th",
  fuels = c("pct_conf", "mean_evi"))
vpd_ext_hp <- gam.hp(m_vpd_ext, iv)
vpd_ext_hp
```

Moisture contributes most to the model, explained 51.9% of the deviance compared to 47.6% for fuels and 0.44% for fire weather
However, fuels may be more important in predicting extreme burn severity compared to median burn severity 

```{r}
iv <- list(
  moisture = c("te(cmi_wy,LL,by=weights_btl2,k=c(3,4),bs=c(\"tp\",\"cr\"))", "te(cmi_wy,LL,by=weights_ln2,k=c(3,4),bs=c(\"tp\",\"cr\"))", "te(cmi_wy,LL,by=weights_lsj2,k=c(3,4),bs=c(\"tp\",\"cr\"))"), 
  fire_weather = "fwi_90th",
  fuels = c("pct_conf", "mean_evi"))
cmi_ext_hp <- gam.hp(m_cmi_ext, iv)
cmi_ext_hp

```

Moisture contributes most to the model, explained 61.4% of the deviance compared to 38.16% for fuels and 0.44% for fire weather


# H2: Are short-term or long-term moisture conditions more important?

Compare RMSE using CV 

Since we fit models with REML estimation, we cant compare models using AIC or ANOVA because the fixed effects structure differs between the models. Therefore, we preformed cross-validation to compare model performance using RMSE as the metric. The better model

## Median

```{r}

compare_models_cv(m_vpd_median, m_cmi_median, moisture_data, response_var = "median")


```

Long-term is slightly better modely but nearly equivalent to short-term moisture conditions

## Extreme

```{r}
compare_models_cv(m_vpd_ext, m_cmi_ext, moisture_data, response_var = "pct_95")
```

Long-term is slightly better modely but nearly equivalent to short-term moisture conditions


# H3: Are relationships consistent between median and extreme?


## VPD
```{r}
summary(m_vpd_median)
summary(m_vpd_ext)

```

## CMI
```{r}
summary(m_cmi_median)
summary(m_cmi_ext)

```


