---
title: "Median DLNM Analysis"
author: "Jack A. Goldman"
date: "2025-12-26"
format: html
---

This use mgcv and dlnm packages to fit a distributed lag model for VPD and fire severity (RBR median)
We use a matrix of VPD values for each fire event (GID) with columns representing lags from 1 to 30 days before fire start date

# Objective
- Short-term increases in VPD (drying conditions) in the days leading up to fire ignition will lead to higher burn severity (RBR median)
- Long-term VPD trends (e.g., seasonal averages) may have less influence on burn severity compared to short-term fluctuations by decreasing fuel abundance

- lagged effects of VPD on burn severity will be non-linear, with immediate effects being stronger than delayed effects
- lagged effects of  CMI on burn severity will be non-linear, with immediate effects being stronger than delayed effects


The goal is to model the effects of short-term and long-term moisture conditions on median and extreme burn severity as having a non-linear, lagged effect and compare whether short-term or long-term moisture conditions better explain burn severity patterns.

# Questions

To achieve this goal, I'll test 4 hypotheses for short and long-term moisture conditions (VPD and CMI (maybe RH and FWI?)) on both median and extreme (95th percentile) burn severity using a series of GAMs with distributed lag non-linear models implemented using the `mgcv`

1.  What is the relative impact of moisture conditions on burn severity?
2.  Are short-term or long-term moisture conditions more important?
3.  Are relationships consistent between median and extreme?
4.  Where are there lagged effects?



```{r}
library(mgcv)
library(gam.hp)
library(tidyverse)
library(here)
conflicted::conflict_prefer("select", "dplyr")
conflicted::conflict_prefer("filter", "dplyr")
```


# Read in data

```{r}
moisture_data <- readRDS(here ("data", "analysis_ready","moisture_lagged_non_scaled_data.rds")) 
```

```{r}
#are there nas in the data
sum(is.na(moisture_data))
moisture_data <- moisture_data |>  
  filter(fire_size_ha > 200) |> 
  filter(ecoregion_lvl %in%c("LN", "BTL", "LSJ"))

moisture_data <- moisture_data[-c(116, 141), ] %>%filter(median != 0) |> filter(pct_95 != 0) |> drop_na()

# find fires in the IQR range for median 
iqr_median <- IQR(moisture_data$median)
q1_median <- quantile(moisture_data$median, 0.25)
q3_median <- quantile(moisture_data$median, 0.75)
lower_bound_median <- q1_median - 1.5 * iqr_median
upper_bound_median <- q3_median + 1.5 * iqr_median
moisture_data_iqr<- moisture_data %>%
  filter(median >= lower_bound_median & median <= upper_bound_median)
```

# Build Full Models



## vpd with distributed lag
```{r}

m_vpd_median <- gam(median ~ 
       te(vpd, L,  by = weights_btl, k = 5, bs = c('tp', 'cr')) + # vpd with lag
       te(vpd, L,  by = weights_ln, k = 5, bs = c('tp', 'cr')) + # vpd with lag
       te(vpd, L,  by = weights_lsj, k = 5, bs = c('tp', 'cr')) + # vpd with lag
         s(pct_conf, by = ecoregion_lvl, bs ="fs", m = 1) +  
         s(mean_evi, by = ecoregion_lvl, bs ="fs",m = 1) + 
         s(fwi_90th, by = ecoregion_lvl, bs = "fs",m = 1) +
         s(fyear,bs = "re") +
        s(ecoregion_lvl, bs = "re") ,
      data = moisture_data_iqr,
      family = gaussian(link = "identity"),
      select = TRUE,
      method = "REML")

summary(m_vpd_median)
```

```{r}
m_vpd_ext <- gam(pct_95 ~ 
       te(vpd, L,  by = weights_btl, k = 5, bs = c('tp', 'cr')) + # vpd with lag
       te(vpd, L,  by = weights_ln, k = 5, bs = c('tp', 'cr')) + # vpd with lag
       te(vpd, L,  by = weights_lsj, k = 5, bs = c('tp', 'cr')) + # vpd with lag
         s(pct_conf, by = ecoregion_lvl, bs =c("tp", "fs"), m = 1) +  
         s(mean_evi, by = ecoregion_lvl, bs =c("tp", "fs"), m = 1) + 
         s(fwi_90th, by = ecoregion_lvl, bs =c("tp", "fs"), m = 1) +
         s(fyear, bs = "re") +
        s(ecoregion_lvl, bs = "re") ,
      data = moisture_data_iqr,
      family = gaussian(link = "identity"),
      select = TRUE,
      method = "REML")

summary(m_vpd_ext)
gam.check(m_vpd_ext)
```

## cmi

## cmi with distributed lag
```{r}

m_cmi_median <- gam(median ~ 
       te(cmi_wy, LL,  by = weights_btl2, k = c(5,5), bs = c('tp', 'cr')) + # vpd with lag
       te(cmi_wy, LL,   by = weights_ln2, k = c(5,5), bs = c('tp', 'cr')) + # vpd with lag
       te(cmi_wy, LL,   by = weights_lsj2, k = c(5,5), bs = c('tp', 'cr')) + # vpd with lag # cmi with 5 year lag
        s(pct_conf, by = ecoregion_lvl, bs =c("tp", "fs"), m = 1) +  
         s(mean_evi, by = ecoregion_lvl, bs =c("tp", "fs"), m = 1) + 
         s(fwi_90th, by = ecoregion_lvl, bs =c("tp", "fs"), m = 1) +
        s(fyear,bs = "re") +
        s(ecoregion_lvl, bs = "re") ,
      data = moisture_data_iqr,
      family = scat(link = "identity"),
      method = "REML")
summary(m_cmi_median)
```

```{r}
m_cmi_ext <- gam(pct_95 ~ 
       te(cmi_wy, LL,  by = weights_btl2, k = 5, bs = c('tp', 'cr')) + # vpd with lag
       te(cmi_wy, LL,   by = weights_ln2, k = 5, bs = c('tp', 'cr')) + # vpd with lag
       te(cmi_wy, LL,   by = weights_lsj2, k = 5, bs = c('tp', 'cr')) + # vpd with lag # cmi with 5 year lag
        s(pct_conf, by = ecoregion_lvl, bs =c("tp", "fs"), m = 1) +  
         s(mean_evi, by = ecoregion_lvl, bs =c("tp", "fs"), m = 1) + 
         s(fwi_90th, by = ecoregion_lvl, bs =c("tp", "fs"), m = 1) +
         s(fyear, bs = "re") +
        s(ecoregion_lvl, bs = "re") ,
      data = moisture_data_iqr,
      family = gaussian(link = "identity"),
      select = TRUE,
      method = "REML")

summary(m_cmi_ext)
```

## save models

```{r}

saveRDS(m_vpd_median, here("analysis", "models","m_vpd_median.rds"))
saveRDS(m_cmi_median, here("analysis", "models","m_cmi_median.rds"))
saveRDS(m_vpd_ext, here("analysis", "models","m_vpd_ext.rds"))
saveRDS(m_cmi_ext, here("analysis", "models","m_cmi_ext.rds"))
```

## assess model fits

vpd
```{r}
gam.check(m_vpd_median)
```


cmi
```{r}
gam.check(m_cmi_median)

```

# H1: What is the relative impact of moisture conditions on burn severity?

checked hierarchical partitioning using gam.hp

## median
```{r}
iv.v <- list(
  moisture = c("te(vpd,L,by=weights_btl,k=c(3,30),bs=c(\"tp\",\"cr\"))", "te(vpd,L,by=weights_ln,k=c(3,30),bs=c(\"tp\",\"cr\"))", "te(vpd,L,by=weights_lsj,k=c(3,30),bs=c(\"tp\",\"cr\"))"), 
  fire_weather = "s(fwi_90th,by=ecoregion_lvl)",
  fuels = c("s(pct_conf, by= ecoregion_lvl)", "s(mean_evi, by = ecoregion_lvl)"))
vpd_median_hp <- gam.hp(m_vpd_median, iv.v)
vpd_median_hp
```

Moisture contributes most to the model, explained 57% of the deviance compared to 39% for fuels and 2.1% for fire weather

```{r}
iv.c <- list(
  moisture = c("te(cmi_wy,LL,by=weights_btl2,k=c(5,5),bs=c(\"tp\",\"cr\"))", "te(cmi_wy,LL,by=weights_ln2,k=c(5,5),bs=c(\"tp\",\"cr\"))", "te(cmi_wy,LL,by=weights_lsj2,k=c(5,5),bs=c(\"tp\",\"cr\"))"), 
  fire_weather = "s(fwi_90th,by=ecoregion_lvl)",
  fuels = c("s(pct_conf, by= ecoregion_lvl)", "s(mean_evi, by = ecoregion_lvl)"))
cmi_median_hp <- gam.hp(m_cmi_median, iv.c)
cmi_median_hp
```

Moisture contributes most to the model, explained 62% of the deviance compared to 36% for fuels and 1.8% for fire weather


## Extreme

```{r}
vpd_ext_hp <- gam.hp(m_vpd_ext, iv.v)
vpd_ext_hp
```

Moisture contributes most to the model, explained 51.9% of the deviance compared to 47.6% for fuels and 0.44% for fire weather
However, fuels may be more important in predicting extreme burn severity compared to median burn severity 

```{r}
cmi_ext_hp <- gam.hp(m_cmi_ext, iv.c)
cmi_ext_hp

```

Moisture contributes most to the model, explained 61.4% of the deviance compared to 38.16% for fuels and 0.44% for fire weather

save results
```{r}
saveRDS( vpd_median_hp, here("analysis", "results","vpd_median_hp.rds"))
saveRDS(cmi_median_hp, here("analysis", "results","cmi_median_hp.rds"))
saveRDS(vpd_ext_hp, here("analysis", "results","vpd_ext_hp.rds"))
saveRDS(cmi_ext_hp, here("analysis", "results","cmi_ext_hp.rds"))
```

# H2: Are short-term or long-term moisture conditions more important?

Compare RMSE using CV 

Since we fit models with REML estimation, we cant compare models using AIC or ANOVA because the fixed effects structure differs between the models. Therefore, we preformed cross-validation to compare model performance using RMSE as the metric. The better model

```{r}
source(here("utils", "cv_functions.r"))
```

## Median

```{r}
library(rsample)
# Assuming moisture_data has a 'fyear' column
cv_folds <- group_vfold_cv(moisture_data, group = fyear, v = 5)  # 5 folds by year

# For each fold, fit VPD and CMI models, predict on test set, compute RMSE
rmse_vpd <- numeric(5)
rmse_cmi <- numeric(5)
for (i in 1:5) {
  train <- cv_folds$splits[[i]] %>% training()
  test <- cv_folds$splits[[i]] %>% testing()
  
  # Fit models on train
  m_vpd <- gam(formula(m_vpd_median), data = train,  family = scat(link = "identity"),
      select = TRUE,
      method = "REML")
  m_cmi <- gam(formula(m_cmi_median), data = train, family = scat(link = "identity"),
      select = TRUE,
      method = "REML")
  
  # Predict on test
  pred_vpd <- predict(m_vpd, newdata = test)
  pred_cmi <- predict(m_cmi, newdata = test)
  
  # RMSE
  rmse_vpd[i] <- sqrt(mean((test$median - pred_vpd)^2, na.rm = TRUE))
  rmse_cmi[i] <- sqrt(mean((test$median - pred_cmi)^2, na.rm = TRUE))
}
rmse_vpd_med <- mean(rmse_vpd)  # Compare averages
rmse_cmi_med <- mean(rmse_cmi)
```

Long-term is slightly better modely but nearly equivalent to short-term moisture conditions

## Extreme

```{r}

# For each fold, fit VPD and CMI models, predict on test set, compute RMSE
rmse_vpd <- numeric(5)
rmse_cmi <- numeric(5)
for (i in 1:5) {
  train <- cv_folds$splits[[i]] %>% training()
  test <- cv_folds$splits[[i]] %>% testing()
  
  # Fit models on train
  m_vpd <- gam(formula(m_vpd_ext), data = train,  family = scat(link = "identity"),
      select = TRUE,
      method = "REML")
  m_cmi <- gam(formula(m_cmi_ext), data = train, family = scat(link = "identity"),
      select = TRUE,
      method = "REML")
  
  # Predict on test
  pred_vpd <- predict(m_vpd, newdata = test)
  pred_cmi <- predict(m_cmi, newdata = test)
  
  # RMSE
  rmse_vpd[i] <- sqrt(mean((test$pct_95 - pred_vpd)^2, na.rm = TRUE))
  rmse_cmi[i] <- sqrt(mean((test$pct_95 - pred_cmi)^2, na.rm = TRUE))
}
rmse_vpd_ext <- mean(rmse_vpd)  # Compare averages
rmse_cmi_ext <- mean(rmse_cmi)
```

Long-term is slightly better modely but nearly equivalent to short-term moisture conditions


```{r}
# make empty dataframe with three columns, response, model, rmse
rmse_results <- data.frame(
  response = character(),
  model = character(),
  rmse = numeric()
)

# add rows for each model and response
rmse_results <- rbind(rmse_results, data.frame(response = "median", model = "vpd", rmse = rmse_vpd_med))
rmse_results <- rbind(rmse_results, data.frame(response = "median", model = "cmi", rmse = rmse_cmi_med))
rmse_results <- rbind(rmse_results, data.frame(response = "extreme", model = "vpd", rmse = rmse_vpd_ext))
rmse_results <- rbind(rmse_results, data.frame(response = "extreme", model = "cmi", rmse = rmse_cmi_ext))

# save results
saveRDS(rmse_results, here("analysis", "results","rmse_comparison_median_extreme.rds"))
```

# H3: Are relationships consistent between median and extreme?

```{r}
source(here("utils", "table_functions.r"))
```

## VPD
```{r}
vpd_med_sum <- make_gam_summary_table(m_vpd_median)
vpd_ext_sum <- make_gam_summary_table(m_vpd_ext)

```

## CMI
```{r}
cmi_med_sum <- make_gam_summary_table(m_cmi_median)
cmi_ext_sum <- make_gam_summary_table(m_cmi_ext)

```

combine gt tables together with med as first two and ext as second two
```{r}
# Combine all data
all_data <- bind_rows(vpd_med_sum, cmi_med_sum, vpd_ext_sum, cmi_ext_sum) 

# as gt
all_data_table <- all_data |> gt()

# Save the table
gtsave(all_data_table, here("tables", "gam_summary_table_median_extreme_v4.docx"))
```


