---
title: "Median DLNM Analysis"
author: "Jack A. Goldman"
date: "2025-12-26"
format: html
---

This use mgcv and dlnm packages to fit a distributed lag model for VPD and fire severity (RBR median)
We use a matrix of VPD values for each fire event (GID) with columns representing lags from 1 to 30 days before fire start date

# Objective
- Short-term increases in VPD (drying conditions) in the days leading up to fire ignition will lead to higher burn severity (RBR median)
- Long-term VPD trends (e.g., seasonal averages) may have less influence on burn severity compared to short-term fluctuations by decreasing fuel abundance

- lagged effects of VPD on burn severity will be non-linear, with immediate effects being stronger than delayed effects
- lagged effects of  CMI on burn severity will be non-linear, with immediate effects being stronger than delayed effects


The goal is to model the effects of short-term and long-term moisture conditions on median and extreme burn severity as having a non-linear, lagged effect and compare whether short-term or long-term moisture conditions better explain burn severity patterns.

# Questions

To achieve this goal, I'll test 4 hypotheses for short and long-term moisture conditions (VPD and CMI (maybe RH and FWI?)) on both median and extreme (95th percentile) burn severity using a series of GAMs with distributed lag non-linear models implemented using the `mgcv`

1.  What is the relative impact of moisture conditions on burn severity?
2.  Are short-term or long-term moisture conditions more important?
3.  Are relationships consistent between median and extreme?
4.  Do lagged effects matter?
5.  Where are there lagged effects?



```{r}
library(mgcv)
library(gam.hp)
library(tidyverse)
library(here)
conflicted::conflict_prefer("select", "dplyr")
conflicted::conflict_prefer("filter", "dplyr")
```


# Read in data

```{r}
moisture_data <- readRDS(here ("data", "analysis_ready","moisture_lagged_non_scaled_data.rds")) 
```

```{r}
#are there nas in the data
sum(is.na(moisture_data))
moisture_data <- moisture_data |> drop_na()
```


# Build Full Distributed Lag


## vpd 
```{r}

m_median <- gam(median~ 
        te(cmi_wy, LL, k = c(30,5))+
       te(vpd, L, k = c(10,30))+
        s(pct_conf, k = 10) +  
         s(dc_90th, k = 10) +
    s(fire_size_class, bs = "re") +
  s(ecoregion_lvl, bs = "re") ,
      data = moisture_data,
      family = scat(link = "identity"),
      method = "REML")

```

```{r}
m_ext <- gam(pct_95~ 
      te(cmi_wy, LL, k = c(30,5))+
       te(vpd, L, k = c(10, 30))+
        s(pct_conf, k = 10) +  
         s(dc_90th, k = 10) +
    s(fire_size_class, bs = "re") +
  s(ecoregion_lvl, bs = "re") ,
      data = moisture_data,
      family = scat(link = "identity"),
      method = "REML")

```



## save models

```{r}

saveRDS(m_median, here("analysis", "models","m_median.rds"))
saveRDS(m_ext, here("analysis", "models","m_ext.rds"))
```

## assess model fits

median
```{r}
gam.check(m_median)
```

ext
```{r}
gam.check(m_median)

```

# H1/H2: What is the relative impact of moisture conditions on burn severity?

checked hierarchical partitioning using gam.hp

## median
```{r}
iv <- list(
  intra = "te(vpd,L,bs=c(\"tp\",\"cr\"))",
  inter = "te(cmi_wy,LL,bs=c(\"tp\",\"cr\"))",
  fire_weather = "s(dc_90th,by=ecoregion_lvl)",
  fuels = "s(pct_conf, by= ecoregion_lvl)")
median_hp <- gam.hp(m_median, iv)
median_hp
```


```{r}
extreme_hp <- gam.hp(m_ext, iv)
extreme_hp
```


### save results
```{r}
saveRDS(median_hp, here("analysis", "results","median_hp.rds"))
saveRDS(extreme_hp, here("analysis", "results","extreme_hp.rds"))

```


# H3: Are relationships consistent between median and extreme?

```{r}
source(here("utils", "table_functions.r"))
```

## VPD
```{r}
med_sum <- make_gam_summary_table(m_median)
ext_sum <- make_gam_summary_table(m_ext)

```


combine gt tables together with med as first two and ext as second two
```{r}
# Combine all data
all_data <- bind_rows(med_sum, ext_sum) 

# as gt
all_data_table <- all_data |> gt()

# Save the table
gtsave(all_data_table, here("tables", "gam_summary_table_median_extreme_v5.docx"))
```

# H4: Do lagged effects matter?

## Median

```{r}
sp_median <- m_median$sp

mm <- gam(median~ 
      cmi_wy_ma_5 + 
    vpd_ma_30 +
        s(pct_conf, k = 10, sp = sp_median["s(pct_conf)"]) +  
         s(dc_90th, k = 10,  sp = sp_median["s(dc_90th)"]) +
  s(ecoregion_lvl, bs = "re",   sp = sp_median["s(ecoregion_lvl)"]) + 
  s(fire_size_class, bs = "re",   sp = sp_median["s(fire_size_class)"]) ,
      data = moisture_data,
      family = scat(link = "identity"),
      method = "REML") 

```

```{r}
summary(mm)
```

Lagged effects more important for median than non-lagged effects 
- lower r2 for non-lagged model

```{r}
h4_lag_median <- anova.gam(m_median, mm, test = "Chisq")
h4_lag_median
```

- lagged effects more important for median than averages

## Extreme

```{r}
sp_ext <- m_ext$sp

me <- gam(pct_95~ 
      cmi_wy_ma_5 + 
    vpd_ma_30 +
        s(pct_conf, k = 10, sp = sp_ext["s(pct_conf)"]) +  
         s(dc_90th, k = 10,  sp = sp_ext["s(dc_90th)"]) +
  s(ecoregion_lvl, bs = "re",   sp = sp_ext["s(ecoregion_lvl)"]) + 
  s(fire_size_class, bs = "re",   sp = sp_ext["s(fire_size_class)"]) ,
      data = moisture_data,
      family = scat(link = "identity"),
      method = "REML") 

```


```{r}
summary(me)
```

Lagged effects of CMI more important 
Mean VPD more important

```{r}
h4_lag_extreme <- anova.gam(m_ext, me, test = "Chisq")
h4_lag_extreme
```

lagged effects more important

```{r}
me2 <- gam(pct_95~ 
      te(cmi_wy, LL, k = c(30,5), sp = sp_ext[c("te(cmi_wy,LL)1", "te(cmi_wy,LL)2")])+
    vpd_ma_30 +
        s(pct_conf, k = 10, sp = sp_ext["s(pct_conf)"]) +  
         s(dc_90th, k = 10,  sp = sp_ext["s(dc_90th)"]) +
  s(ecoregion_lvl, bs = "re",   sp = sp_ext["s(ecoregion_lvl)"]) + 
  s(fire_size_class, bs = "re",   sp = sp_ext["s(fire_size_class)"]) ,
      data = moisture_data,
      family = scat(link = "identity"),
      method = "REML") 

summary(me2)

anova.gam(me, me2, test = "Chisq")

iv3<- list(
  intra = "vpd_ma_30",
  inter = "te(cmi_wy,LL)",
  fire_weather = "s(dc_90th)",
  fuels = "s(pct_conf)")
gam.hp::gam.hp(me2, iv3)


```

