---
title: "Exploratory Data Analysis and Descriptive Statistics"
author: "Jack A. Goldman"
date: "2025-12-26"
format: html
---


```{r}
library(ggplot2)
library(tidyr)
library(ggplot2)
library(viridis)
library(cowplot)
library(patchwork)
library(readr)
library(mgcv)
library(dlnm)
library(reshape2)
library(GGally)
library(dplyr)
library(here)
library(ggridges)
```


# Read in data

```{r}
moisture_data <- readRDS(here("data", "analysis_ready", "moisture_lagged_data.rds")) |> 
  filter(fire_size_ha > 200) |> 
  filter(ecoregion_lvl %in%c("LN", "BTL", "LSJ"))


moisture_df <- moisture_df |>  
  filter(fire_size_ha > 200) |> 
  filter(ecoregion_lvl %in%c("LN", "BTL", "LSJ"))

moisture_df <- moisture_df[-c(116, 141), ] %>%filter(median != 0) |> filter(pct_95 != 0)

moisture_df <- moisture_df |> select(-c(mean_ndmi, mean_ndvi)) |>  drop_na()
```

# ecoregion levels
```{r}
unique(moisture_data$ecoregion)
# counts per ecoregion 
ecoregion_counts <- moisture_data |> 
    group_by(ecoregion_lvl) |> 
    summarise(count = n())

print(ecoregion_counts)
```

# fires per year

```{r}
fires_per_year <- moisture_data |> 
    group_by(fyear) |> 
    summarise(num_fires = n())
print(fires_per_year, n = Inf)

ggplot(fires_per_year, aes(x = fyear, y = num_fires)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    theme_minimal() +
    labs(title = "Number of Fires per Year", x = "Year", y = "Number of Fires")

# geom ridgline plot of median by year
Use theme(legend.position = "none") and limit y-axis breaks to the four years. Replace the selection with:



m_ridge <- {
    yrs <- sort(unique(as.character(moisture_data$fyear)))
    ggplot(moisture_data, aes(x = median, y = as.factor(fyear), fill = ..x..)) +
        ggridges::geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
        scale_fill_viridis_c(option = "C") +
        scale_y_discrete(
            breaks = yrs,
            labels = function(x) ifelse(x %in% c("1990", "2000", "2010", "2020"), x, "")
        ) +
        labs(x = "Median", y = "Year") +
        theme_bw() +
        theme(legend.position = "none")
}

q95_ridge <- {
    yrs <- sort(unique(as.character(moisture_data$fyear)))
    ggplot(moisture_data, aes(x = pct_95, y = as.factor(fyear), fill = ..x..)) +
        ggridges::geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
        scale_fill_viridis_c(option = "C") +
        scale_y_discrete(
            breaks = yrs,
            labels = function(x) ifelse(x %in% c("1990", "2000", "2010", "2020"), x, "")
        ) +
        labs(x = "Extreme Burn Severity", y = "Year") +
        theme_bw() +
        theme(legend.position = "none")
}



# create  table of fires per year using gt package

fy_tbl <- fires_per_year |> 
    rename(Year = fyear, `Number of Fires` = num_fires) |> 
    gt() |>
    tab_options(
        table.font.names = "Arial",
        table.font.size = px(8)
    ) 
           


# patchwork group ridgeline plts and table
library(patchwork)
p1 <- (m_ridge | q95_ridge) +
    plot_layout(heights = c(3, 1)) + plot_annotation(tag_levels = 'A') 



ggsave(here("figures", "fires_per_year_ridgeline.png"), width = 6, height = 4)

```

# variantion in severity across years and ecoregions

```{r}
year_eco <- moisture_data |> 
    select(c(year, ecoregion_lvl, median, pct_95, fire_size_ha, mean_evi)) |> 
    drop_na() |> 
    select(-c(mean_evi))

#  total hectares burned per year and ecoregion and total
burned_area_summary <- year_eco |> 
    group_by(year, ecoregion_lvl) |> 
    summarise(total_burned_ha = sum(fire_size_ha),
              mean_median = mean(median),
              mean_pct_95 = mean(pct_95),
              num_fires = n()) |> 
    ungroup()

# total burned area
total_area_burned <- sum(year_eco$fire_size_ha)
total_fires <- 667

# total ecoregion
ecoregion_summary <- year_eco |> 
    group_by(ecoregion_lvl) |> 
    summarise(total_burned_ha = sum(fire_size_ha),
              mean_median = mean(median),
              mean_pct_95 = mean(pct_95),
              num_fires = n()) |> 
    ungroup()
# ecoregion percent of total
ecoregion_summary <- ecoregion_summary |> 
    mutate(percent_total = (total_burned_ha / total_area_burned) * 100,
    percent_fires = (num_fires / total_fires) * 100)



```


# trend over time
```{r}
library(trend)
# Aggregate to annual means by ecoregion and year
annual_data <- moisture_data %>%
  mutate(year = as.numeric(as.character(fyear))) %>%  # Convert fyear factor to numeric
  group_by(ecoregion_lvl, year) %>%
  summarize(
    mean_median = mean(median, na.rm = TRUE),
    mean_pct_95 = mean(pct_95, na.rm = TRUE),
    .groups = "drop"
  )

# Calculate Sen's slope and p-values for each ecoregion
trends <- annual_data %>%
  group_by(ecoregion_lvl) %>%
  summarize(
    sen_median_slope = sens.slope(mean_median[order(year)])$estimates["Sen's slope"],
    sen_median_p = sens.slope(mean_median[order(year)])$p.value,
    sen_pct_95_slope = sens.slope(mean_pct_95[order(year)])$estimates["Sen's slope"],
    sen_pct_95_p = sens.slope(mean_pct_95[order(year)])$p.value
  )

# View the results
trends

# lm trends
lm_trends <- annual_data %>%
  group_by(ecoregion_lvl) %>%
  summarize(
    lm_median_slope = coef(lm(mean_median ~ year))[2],
    lm_median_p = summary(lm(mean_median ~ year))$coefficients[2, 4],
    lm_pct_95_slope = coef(lm(mean_pct_95 ~ year))[2],
    lm_pct_95_p = summary(lm(mean_pct_95 ~ year))$coefficients[2, 4]
  )
```

Stacked bar plot with fire year on x and count on y with and colour = ecoregions
```{r}



fy_plot <- ggplot(moisture_data, aes(x = as.integer(as.character(fyear)), fill = ecoregion_lvl)) +
    geom_bar(position = "stack") +
    scale_x_continuous(
        breaks = c(1990, 1995, 2000, 2005, 2010, 2015, 2020),
        labels = function(x) ifelse(x %in% c(1990, 2000, 2010, 2020), as.character(x), "")
    ) +
    scale_fill_manual(
        values = c("BTL" = "gray60", "LN" = "darkorange", "LSJ" = "forestgreen"),
        name = "Ecoregion"
    ) +
    labs(x = "Year", y = "Number of Fires") +
    theme_bw()



# years summary of mean median and pct_95 
yearly_summary <- moisture_data %>%
  group_by(fyear) %>%
  summarize(
    mean_median = mean(median, na.rm = TRUE),
    mean_pct_95 = mean(pct_95, na.rm = TRUE)
  ) |> arrange(desc(mean_pct_95))

# make a boxplot of mean median and pct_95 by year with ecoregion as colour 
med_y <- ggplot(moisture_data, aes(x = as.factor(fyear), y = median, fill = ecoregion_lvl)) +
    geom_boxplot() +
    scale_x_discrete(
        breaks = c("1990", "1995", "2000", "2005", "2010", "2015", "2020"),
        labels = function(x) ifelse(x %in% c("1990", "2000", "2010", "2020"), x, "")
    ) +
    scale_fill_manual(
        values = c("BTL" = "gray60", "LN" = "darkorange", "LSJ" = "forestgreen"),
        name = "Ecoregion"
    ) +
    labs(x = "Year", y = "Median Burn Severity") +
    theme_bw()

ext_y <- ggplot(moisture_data, aes(x = as.factor(fyear), y = pct_95, fill = ecoregion_lvl)) +
    geom_boxplot() +
    scale_x_discrete(
        breaks = c("1990", "1995", "2000", "2005", "2010", "2015", "2020"),
        labels = function(x) ifelse(x %in% c("1990", "2000", "2010", "2020"), x, "")
    ) +
    scale_fill_manual(
        values = c("BTL" = "gray60", "LN" = "darkorange", "LSJ" = "forestgreen"),
        name = "Ecoregion"
    ) +
    labs(x = "Year", y = "Extreme Burn Severity") +
    theme_bw()



# combine them with patchwork in ncol = 3
# collect guides and move below plots



p2 <- (fy_plot| med_y | ext_y) + plot_annotation(tag_levels = 'A') + 
    plot_layout(guides = 'collect') & theme(legend.position = 'bottom')

p3 <- (med_y / ext_y) + plot_annotation(tag_levels = 'A') + 
    plot_layout(guides = 'collect') & theme(legend.position = 'bottom')

fy_plot2 <- fy_plot + guides(color = "none", fill = "none", linetype = "none", shape = "none", size = "none")

med_y2 <- med_y +  guides(color = "none", fill = "none", linetype = "none", shape = "none", size = "none")

p4 <- (fy_plot2 + med_y2 + ext_y) + guide_area() + plot_layout(guides = 'collect' )+ plot_annotation(tag_levels = 'A') 

p5 <- (fy_plot2 | med_y2 / ext_y) + plot_layout(guides = 'collect' )+ theme(legend.position = "bottom")+ plot_annotation(tag_levels = 'A') 

# save it ggsave(here("figures", "fires_per_year_stacked.png"), width = 6, height = 4)
ggsave(here("figures", "fires_per_year_median_ext_eco.png"), width = 10, height = 6)

```

list of fires per year
```{r}
fires_per_year_ecoregion <- moisture_data %>%
  group_by(fyear, ecoregion_lvl) %>%
  summarize(num_fires = n()) %>%
  arrange(fyear, ecoregion_lvl) |> arrange(desc(num_fires))

fires_per_year <- moisture_data %>%
  group_by(fyear) %>%
  summarize(num_fires = n()) %>%
  arrange(desc(num_fires))

```

 # fire size trends per ecoregion

 ```{r}
fire_size_trends <- moisture_data %>%
  mutate(year = as.numeric(as.character(fyear))) %>%
  group_by(ecoregion_lvl) %>%
  summarize(
    lm_fire_size_slope = coef(lm(fire_size_ha ~ year))[2],
    lm_fire_size_p = summary(lm(fire_size_ha ~ year))$coefficients[2, 4]
  )
 ```


# forest cover percent per ecoregion

```{r}
forest_cover <- moisture_df |> #non-scaled df
    group_by(ecoregion_lvl) |> 
    summarise(mean_conf_pct = mean(pct_conf, na.rm = TRUE),
    mean_up_pct = mean(pct_up, na.rm = TRUE),
    mean_wetland = mean(pct_wet, na.rm = TRUE))

# variation in conifer percent per ecoregion
forest_cover_sd <- moisture_df |> #non-scaled df
    group_by(ecoregion_lvl) |> 
    summarise(sd_conf_pct = sd(pct_conf, na.rm = TRUE),
    sd_up_pct = sd(pct_up, na.rm = TRUE),
    sd_wetland = sd(pct_wet, na.rm = TRUE))

# mean evi per ecoregion
evi_ecoregion <- moisture_df |> #non-scaled df
    group_by(ecoregion_lvl) |> 
    summarise(mean_evi = mean(mean_evi, na.rm = TRUE),
    sd_evi = sd(mean_evi, na.rm = TRUE))

# scatter plot of evi by burn severity
ggplot(moisture_df, aes(x = mean_evi, y = median, color = ecoregion_lvl)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", se = FALSE) +
    labs(x = "Mean EVI", y = "Median Burn Severity", color = "Ecoregion") +
    theme_minimal()

```

# get different quantiles of pct_conf from moisture_df as dataframe
```{r}
#column to moisture_df classifying pct_conf into low, medium, high based on quantiles
moisture_df <- moisture_df |> 
  mutate(pct_conf_class = cut(pct_conf, 
                              breaks = quantile(pct_conf, probs = c(0, 0.33, 0.67, 1), na.rm = TRUE),
                              labels = c("Low", "Medium", "High"),
                              include.lowest = TRUE))

# correlation between pct_conf and median and pct_95 and different quantiles
low_con <- moisture_df |> filter(pct_conf_class == "Low")

cor(low_con$pct_conf, low_con$median, use="complete.obs")

med_con <- moisture_df |> filter(pct_conf_class == "Medium")
cor(med_con$pct_conf, med_con$median, use="complete.obs")

high_con <- moisture_df |> filter(pct_conf_class == "High")
cor(high_con$pct_conf, high_con$median, use="complete.obs")

# relationship across quantiles
ggplot(moisture_df, aes(x = pct_conf, y = median, color = pct_conf_class)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", se = FALSE) +
    labs(x = "Percent Conifer Cover", y = "Median Burn Severity", color = "Conifer Cover Class") +
    theme_minimal()

ggplot(moisture_df, aes(x = pct_conf_class, y = median)) +
  geom_boxplot() +
  labs(x = "Conifer Class", y = "Median Burn Severity") +
  theme_minimal()

#ANOVA for severity differences by class
aov_result <- aov(median ~ pct_conf_class, data = moisture_df)
summary(aov_result)

# Correlations by class (as in your session)
low_con <- moisture_df %>% filter(pct_conf_class == "Low")
med_con <- moisture_df %>% filter(pct_conf_class == "Medium")
high_con <- moisture_df %>% filter(pct_conf_class == "High")

cor(low_con$pct_conf, low_con$median, use = "complete.obs")  # Low class - very weak negative correlation, high conifer cover is slightly associated with lower severity  - negligible effect
cor(med_con$pct_conf, med_con$median, use = "complete.obs")  # Medium - almost no relationship.
cor(high_con$pct_conf, high_con$median, use = "complete.obs")  # High - Higher conifer cover is associated with higher severity.
# This suggests a threshold effect. In low/medium conifer areas, pct_conf has little to no impact on severity (possibly due to other limiting factors like moisture). But in high-conifer areas (>~67% quantile), it becomes a driver, increasing severity—aligning with your hypotheses on fuel availability.


# indentify threshold effects of pct_conf
library(segmented)
# fit linear model
# Simple linear model first
lm_model <- lm(median ~ pct_conf, data = moisture_data)

# Fit segmented model (allows 1 breakpoint)
seg_model <- segmented(lm_model, seg.Z = ~ pct_conf, npsi = 1)
summary(seg_model)

# Plot to visualize breakpoint
plot(seg_model, res = TRUE, col = "red")  # Residuals and fit
```



# range of median and pct_95 

```{r}
summary(moisture_data$median)
summary(moisture_data$pct_95)
sd(moisture_data$median)
sd(moisture_data$pct_95)
```

# range of model response/predictors as well as sd as gt table
```{r}
vars <- moisture_data |> 
    select(cmi_wy, vpd, pct_conf, fwi_90th, mean_evi)

# back transform all vars (scaled mean and sd*2)


```


```{r}



cor(moisture_data$pct_up, moisture_data$cmi_wy, use="complete.obs")
cor(moisture_data$pct_up, moisture_data$vpd,    use="complete.obs")


summary(moisture_df$pct_up); summary(moisture_df$pct_up)
range(moisture_df$pct_up);  range(moisture_df$pct_up)


```

# correlation between continuous variables except vpd and L

```{r}
colnames(moisture_data)
cor_matrix <- moisture_data |> 
    select(c(2:21)) |> # remove first column and 15 to 24th column
    #select only numeric columns
    select(where(is.numeric)) |>
    cor()

# look at correlation between ndmi, evi and ndmi columns
cor_moist <- moisture_data |> 
    select(c(mean_ndmi, mean_evi, mean_ndvi)) |> drop_na() |> 
    cor()
# apply a heatmap to visualize the correlation matrix

cor_melt <- melt(cor_matrix)
ggplot(data = cor_melt, aes(x=Var1, y=Var2, fill=value)) +
    geom_tile() +
    scale_fill_viridis(option = "C", limits = c(-1, 1)) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
    labs(title = "Correlation Matrix Heatmap", x = "", y = "", fill = "Correlation")

# find instances where correlation is > 0.5 or < -0.5
high_cor <- cor_melt |> 
    filter(value > 0.5 | value < -0.5 & Var1 != Var2)
high_cor |> filter(Var1 == "pct_up" | Var2 == "pct_up")

```


```{r}
ggpairs(moisture_data |>  select(c(-1, -15:-24)) |> # remove first column and 15 to 24th column
    #select only numeric columns
    select(where(is.numeric)) )
```

# out of all variables with grep(90th), plot the relationship with median and pct_90 facet wrap by varable name

```{r}
moisture_data_long <- moisture_data |> 
    pivot_longer(cols = contains("90th"), names_to = "variable", values_to = "value")

ggplot(moisture_data_long, aes(x = value, y = median)) +
    geom_point(alpha = 0.5) +
    theme_minimal() +
    facet_wrap(~ variable, scales = "free_x") +
    labs(title = "Relationship with FWI duration", x = "Value", y = "Median") +
    scale_color_viridis(discrete = TRUE)

```


Get number of fires per year and make a histogram


```{r}
fires_per_year <- moisture_data |> 
    group_by(year) |> 
    summarise(num_fires = n())

print(fires_per_year, n = Inf)

ggplot(fires_per_year, aes(x = year, y = num_fires)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    theme_minimal() +
    labs(title = "Number of Fires per Year", x = "Year", y = "Number of Fires")
```

What is the fire size in year == 2022
```{r}
year_2022 <- moisture_data |> 
    filter(year == 2022)
summary(year_2022$fire_size)

```

how many fires per season

```{r}
seaosn_summary <- moisture_data |> 
    group_by(season) |> 
    summarise(num_fires = n())
```

Because it is heavily skewed to summer fires (1032 vs 56), we will just include summer fires (after july 1)


# fire size distribution 
```{r}
# how many fires with fire_size_ha > 200
mini_fires <- moisture_data |> 
    filter(fire_size_ha < 200)
nrow(mini_fires)

# small (as per stocks)
small_fires <- moisture_data |> 
    filter(fire_size_ha >= 200 & fire_size_ha < 500)
nrow(small_fires)

# medium (as per stocks)
medium_fires <- moisture_data |> 
    filter(fire_size_ha >= 500 & fire_size_ha < 10000)
nrow(medium_fires)

# large (as per stocks)
large_fires <- moisture_data |> 
    filter(fire_size_ha >= 10000)
nrow(large_fires)

```


```{r}
#plot median by mean evi using a loess smoother
ggplot(moisture_data, aes(x = mean_ndmi, y = pct_90)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "loess", color = "blue") +
    theme_minimal() +
    labs(title = "Median vs Mean EVI", x = "Mean EVI", y = "Median")
```


