---
title: "Exploratory Data Analysis and Descriptive Statistics"
author: "Jack A. Goldman"
date: "2025-12-26"
format: html
---


```{r}
library(ggplot2)
library(tidyr)
library(ggplot2)
library(viridis)
library(cowplot)
library(patchwork)
library(readr)
library(mgcv)
library(dlnm)
library(reshape2)
library(GGally)
library(dplyr)
library(here)
library(ggridges)
```


# Read in data

```{r}
moisture_data <- readRDS(here("data", "analysis_ready", "moisture_lagged_non_scaled_data.rds")) |>   drop_na()
```

# Fire Size

```{r}
sum(moisture_data$fire_size_ha)
```

fire size by ecoregion
```{r}
moisture_data |> 
    group_by(ecoregion_lvl) |> 
    summarise(total_fire_size_ha = sum(fire_size_ha))

#get percent of total per ecoregion
moisture_data |> 
    group_by(ecoregion_lvl) |> 
    summarise(total_fire_size_ha = sum(fire_size_ha)) |> 
    mutate(percent_total = (total_fire_size_ha / sum(total_fire_size_ha)) * 100) 
```

fire size class distribution
```{r}
fire_size_class_counts <- moisture_data |> 
    group_by(fire_size_class) |> 
    summarise(count = n()) |> 
    mutate(percent_total = (count / sum(count)) * 100)
print(fire_size_class_counts)
ggplot(fire_size_class_counts, aes(x = fire_size_class, y = count)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    theme_minimal() +
    labs(title = "Fire Size Class Distribution", x = "Fire Size Class", y = "Count of Fires")
```

# ecoregion levels
```{r}
unique(moisture_data$ecoregion)
# counts per ecoregion 
ecoregion_counts <- moisture_data |> 
    group_by(ecoregion_lvl) |> 
    summarise(count = n())

print(ecoregion_counts)
```



# fires per year

```{r}
fires_per_year <- moisture_data |> 
    group_by(fyear) |> 
    summarise(num_fires = n()) |> arrange(desc(num_fires))
print(fires_per_year, n = Inf) 

ggplot(fires_per_year, aes(x = fyear, y = num_fires)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    theme_minimal() +
    labs(title = "Number of Fires per Year", x = "Year", y = "Number of Fires")

# geom ridgline plot of median by year
Use theme(legend.position = "none") and limit y-axis breaks to the four years. Replace the selection with:



m_ridge <- {
    yrs <- sort(unique(as.character(moisture_data$fyear)))
    ggplot(moisture_data, aes(x = median, y = as.factor(fyear), fill = ..x..)) +
        ggridges::geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
        scale_fill_viridis_c(option = "C") +
        scale_y_discrete(
            breaks = yrs,
            labels = function(x) ifelse(x %in% c("1990", "2000", "2010", "2020"), x, "")
        ) +
        labs(x = "Median", y = "Year") +
        theme_bw() +
        theme(legend.position = "none")
}

q95_ridge <- {
    yrs <- sort(unique(as.character(moisture_data$fyear)))
    ggplot(moisture_data, aes(x = pct_95, y = as.factor(fyear), fill = ..x..)) +
        ggridges::geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
        scale_fill_viridis_c(option = "C") +
        scale_y_discrete(
            breaks = yrs,
            labels = function(x) ifelse(x %in% c("1990", "2000", "2010", "2020"), x, "")
        ) +
        labs(x = "Extreme Burn Severity", y = "Year") +
        theme_bw() +
        theme(legend.position = "none")
}





# patchwork group ridgeline plts and table
library(patchwork)
p1 <- (m_ridge | q95_ridge) +
    plot_layout(heights = c(3, 1)) + plot_annotation(tag_levels = 'A') 



ggsave(here("figures", "fires_per_year_ridgeline.png"), width = 6, height = 4)

```

# variantion in severity across years and ecoregions

```{r}
year_eco <- moisture_data |> 
    select(c(year, ecoregion_lvl, median, pct_95, fire_size_ha)) |> 
    drop_na() 

#  total hectares burned per year and ecoregion and total
burned_area_summary <- year_eco |> 
    group_by(year, ecoregion_lvl) |> 
    summarise(total_burned_ha = sum(fire_size_ha),
              mean_median = mean(median),
              mean_pct_95 = mean(pct_95),
              num_fires = n()) |> 
    ungroup()

# total burned area
total_area_burned <- sum(year_eco$fire_size_ha)
total_fires <- 667

# total ecoregion
ecoregion_summary <- year_eco |> 
    group_by(ecoregion_lvl) |> 
    summarise(total_burned_ha = sum(fire_size_ha),
              mean_median = mean(median),
              mean_pct_95 = mean(pct_95),
              num_fires = n()) |> 
    ungroup()
# ecoregion percent of total
ecoregion_summary <- ecoregion_summary |> 
    mutate(percent_total = (total_burned_ha / total_area_burned) * 100,
    percent_fires = (num_fires / total_fires) * 100)



```



Stacked bar plot with fire year on x and count on y with and colour = ecoregions
```{r}

eco_colours <- c("BTL" = "gray30", "LN" = "orangered", "LSJ" = "firebrick", "LW" = "darkgreen", "PR" = "goldenrod", "ACP" = "deepskyblue4")


fy_plot <- ggplot(moisture_data, aes(x = as.integer(as.character(fyear)), fill = ecoregion_lvl)) +
    geom_bar(position = "stack") +
    scale_x_continuous(
        breaks = c(1990, 1995, 2000, 2005, 2010, 2015, 2020),
        labels = function(x) ifelse(x %in% c(1990, 2000, 2010, 2020), as.character(x), "")
    ) +
    scale_fill_manual(
        values = eco_colours,
        name = "Ecoregion"
    ) +
    labs(x = "Year", y = "Number of Fires") +
    theme_bw()



# years summary of mean median and pct_95 
yearly_summary <- moisture_data %>%
  group_by(fyear) %>%
  summarize(
    mean_median = mean(median, na.rm = TRUE),
    mean_pct_95 = mean(pct_95, na.rm = TRUE)
  ) |> arrange(desc(mean_pct_95))

# make a boxplot of mean median and pct_95 by year with ecoregion as colour 
Add fill = "firebrick3" to geom_boxplot. Modified selection:


med_y <- ggplot(moisture_data, aes(x = as.factor(fyear), y = median)) +
    geom_boxplot(fill = "firebrick3") +
    scale_x_discrete(
        breaks = c("1990", "1995", "2000", "2005", "2010", "2015", "2020"),
        labels = function(x) ifelse(x %in% c("1990", "2000", "2010", "2020"), x, "")
    )  +
    labs(x = "Year", y = "Median Burn Severity") +
    theme_bw()


ext_y <- ggplot(moisture_data, aes(x = as.factor(fyear), y = pct_95)) +
    geom_boxplot(fill = "firebrick3") +
    scale_x_discrete(
        breaks = c("1990", "1995", "2000", "2005", "2010", "2015", "2020"),
        labels = function(x) ifelse(x %in% c("1990", "2000", "2010", "2020"), x, "")
    ) +
    labs(x = "Year", y = "Extreme Burn Severity") +
    theme_bw()



# combine them with patchwork in ncol = 3
# collect guides and move below plots





fy_plot2 <- fy_plot + theme(
    legend.position = c(0.98, 0.97),       # Position legend box at top-right of panel
    legend.justification = c(1, 1), # Anchor the legend's top-right corner to that position
    legend.background = element_rect(fill = "white", color = NA), # Optional: Add a white background without border
    legend.margin = margin(t = 5, r = 5, b = 5, l = 5) ,
    legend.key.size = unit(0.5, "cm")# Optional: Adjust margins around the legend content
  ) 

ea() + plot_layout(guides = 'collect' )+ plot_annotation(tag_levels = 'A') 

p5 <- (fy_plot2 | med_y / ext_y) + plot_annotation(tag_levels = 'A') 

# save it ggsave(here("figures", "fires_per_year_stacked.png"), width = 6, height = 4)
ggsave(here("figures", "fires_per_year_median_ext_eco.png"), width = 10, height = 6)

```

list of fires per year
```{r}
fires_per_year_ecoregion <- moisture_data %>%
  group_by(fyear, ecoregion_lvl) %>%
  summarize(num_fires = n()) %>%
  arrange(fyear, ecoregion_lvl) |> arrange(desc(num_fires))

fires_per_year <- moisture_data %>%
  group_by(fyear) %>%
  summarize(num_fires = n()) %>%
  arrange(desc(num_fires))

```



# forest cover percent per ecoregion

```{r}
moisture_data |> #non-scaled df
    group_by(ecoregion_lvl) |> 
    summarise(mean_conf_pct = mean(pct_conf, na.rm = TRUE))

# variation in conifer percent per ecoregion
 moisture_data |> #non-scaled df
    group_by(ecoregion_lvl) |> 
    summarise(sd_conf_pct = sd(pct_conf, na.rm = TRUE))


# scatter plot of evi by burn severity
ggplot(moisture_data, aes(x = pct_conf, y = median, color = ecoregion_lvl)) +
    geom_point(alpha = 0.5) +
    labs(x = "Percent Conifer Cover", y = "Median Burn Severity", color = "Ecoregion") +
    theme_minimal()

# scatter plot of evi by burn severity
ggplot(moisture_data, aes(x = pct_conf, y = pct_95, color = ecoregion_lvl)) +
    geom_point(alpha = 0.5) +
    labs(x = "Percent Conifer Cover", y = "Extreme Burn Severity", color = "Ecoregion") +
    theme_minimal()

```

```{r}
moisture_data |> 
    group_by(ecoregion_lvl) |> summarise(mean_dc = mean(dc_90th))

# scatter plot of evi by burn severity
ggplot(moisture_data, aes(x = dc_90th, y = median, color = ecoregion_lvl)) +
    geom_point(alpha = 0.5) +
    labs(x = "Drought Code", y = "Median Burn Severity", color = "Ecoregion") +
    theme_minimal()

# scatter plot of evi by burn severity
ggplot(moisture_data, aes(x = dc_90th, y = pct_95, color = ecoregion_lvl)) +
    geom_point(alpha = 0.5) +
    labs(x = "Drought Code", y = "Extreme Burn Severity", color = "Ecoregion") +
    theme_minimal()
```


# get different quantiles of pct_conf from moisture_df as dataframe
```{r}
#column to moisture_df classifying pct_conf into low, medium, high based on quantiles
moisture_df <- moisture_df |> 
  mutate(pct_conf_class = cut(pct_conf, 
                              breaks = quantile(pct_conf, probs = c(0, 0.33, 0.67, 1), na.rm = TRUE),
                              labels = c("Low", "Medium", "High"),
                              include.lowest = TRUE))

# correlation between pct_conf and median and pct_95 and different quantiles
low_con <- moisture_df |> filter(pct_conf_class == "Low")

cor(low_con$pct_conf, low_con$median, use="complete.obs")

med_con <- moisture_df |> filter(pct_conf_class == "Medium")
cor(med_con$pct_conf, med_con$median, use="complete.obs")

high_con <- moisture_df |> filter(pct_conf_class == "High")
cor(high_con$pct_conf, high_con$median, use="complete.obs")

# relationship across quantiles
ggplot(moisture_df, aes(x = pct_conf, y = median, color = pct_conf_class)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", se = FALSE) +
    labs(x = "Percent Conifer Cover", y = "Median Burn Severity", color = "Conifer Cover Class") +
    theme_minimal()

ggplot(moisture_df, aes(x = pct_conf_class, y = median)) +
  geom_boxplot() +
  labs(x = "Conifer Class", y = "Median Burn Severity") +
  theme_minimal()

#ANOVA for severity differences by class
aov_result <- aov(median ~ pct_conf_class, data = moisture_df)
summary(aov_result)

# Correlations by class (as in your session)
low_con <- moisture_df %>% filter(pct_conf_class == "Low")
med_con <- moisture_df %>% filter(pct_conf_class == "Medium")
high_con <- moisture_df %>% filter(pct_conf_class == "High")

cor(low_con$pct_conf, low_con$median, use = "complete.obs")  # Low class - very weak negative correlation, high conifer cover is slightly associated with lower severity  - negligible effect
cor(med_con$pct_conf, med_con$median, use = "complete.obs")  # Medium - almost no relationship.
cor(high_con$pct_conf, high_con$median, use = "complete.obs")  # High - Higher conifer cover is associated with higher severity.
# This suggests a threshold effect. In low/medium conifer areas, pct_conf has little to no impact on severity (possibly due to other limiting factors like moisture). But in high-conifer areas (>~67% quantile), it becomes a driver, increasing severity—aligning with your hypotheses on fuel availability.


# indentify threshold effects of pct_conf
library(segmented)
# fit linear model
# Simple linear model first
lm_model <- lm(median ~ pct_conf, data = moisture_data)

# Fit segmented model (allows 1 breakpoint)
seg_model <- segmented(lm_model, seg.Z = ~ pct_conf, npsi = 1)
summary(seg_model)

# Plot to visualize breakpoint
plot(seg_model, res = TRUE, col = "red")  # Residuals and fit
```



# range of median and pct_95 

```{r}
summary(moisture_data$median)
summary(moisture_data$pct_95)
sd(moisture_data$median)
sd(moisture_data$pct_95)
```

# range of model response/predictors as well as sd as gt table
```{r}
vars <- moisture_data |> 
    select(cmi_wy, vpd, pct_conf, fwi_90th, mean_evi)

# back transform all vars (scaled mean and sd*2)


```


```{r}



cor(moisture_data$pct_up, moisture_data$cmi_wy, use="complete.obs")
cor(moisture_data$pct_up, moisture_data$vpd,    use="complete.obs")


summary(moisture_df$pct_up); summary(moisture_df$pct_up)
range(moisture_df$pct_up);  range(moisture_df$pct_up)


```


# trend in cmi and vpd over time

get cmi wy

```{r}
cmi <- read_csv(here("data",  "cmi_hogg_water_year_anomaly.csv"))

# grom GID in cmi, take every before the _ and make new column year
cmi <- cmi |> 
    dplyr::mutate(year = as.integer(sub("_.*", "", GID)))

#plot avg_cmi_anomaly by water year
ggplot(cmi, aes(x = water_year, y = avg_cmi_anomaly)) +
    geom_line(color = "blue") +
    geom_point() +
    theme_minimal() +
    labs(title = "Average CMI Anomaly by Water Year", x = "Water Year", y = "Average CMI Anomaly")

# get trend using gam
gam_cmi <- gam(avg_cmi_anomaly ~ s(water_year, bs = "cc", k = 37), data = cmi, method = "REML")
summary(gam_cmi)

# plot the gam fit
plot(gam_cmi, shade = TRUE, main = "GAM Fit for Average CMI Anomaly over Water Years") 


# how do i get trend from gam
library(mgcv)
library(gratia)
ctrl <- list(niterEM = 0, msVerbose = TRUE, optimMethod="L-BFGS-B")

m2 <- gamm(avg_cmi_anomaly ~ s(water_year, bs = "cc", k = 37) ,
           data = cmi, correlation = corARMA(form = ~ 1|year, p = 2),
           control = ctrl)
summary(m2$lme)
plot(m2$gam, shade = TRUE, main = "GAMM Fit for Average CMI Anomaly over Water Years")

# average cmi per year
df_cmi <- cmi |> 
    group_by(water_year) |> 
    summarise(cmi_anom = mean(avg_cmi_anomaly, na.rm = TRUE))
# average rbr per year
df_sev <- rbr |> 
    group_by(year) |> 
    summarise(mean_rbr = mean(median, na.rm = TRUE))

# as xts
df_cmi_xts <- xts::xts(df_cmi$cmi_anom, order.by = as.Date(paste0(df_cmi$water_year, "-10-01")))
df_sev_xts <- xts::xts(df_sev$mean_rbr, order.by = as.Date(paste0(df_sev$year, "-01-01")))

xts::plot.xts(df_cmi_xts, main = "Average CMI Anomaly Over Water Years", ylab = "CMI Anomaly", xlab = "Year", col = "blue")
xts::plot.xts(df_sev_xts, main = "Average Median Burn Severity Over Years", ylab = "Median Burn Severity", xlab = "Year", col = "red")

# CMI anomaly and median burn severity together


# can you plot both on same plot with two y axes and x axis of water_year
library(ggplot2)
df_combined <- df_cmi %>%
  inner_join(df_sev, by = c("water_year" = "year"))
ggplot(df_combined, aes(x = water_year)) + 
  geom_line(aes(y = cmi_anom, color = "CMI Anomaly")) +
  geom_line(aes(y = mean_rbr, color = "Median Burn Severity")) +
  scale_y_continuous(
    name = "CMI Anomaly",
    sec.axis = sec_axis(~ ., name = "Median Burn Severity")
  ) +
  labs(x = "Year", title = "Average CMI Anomaly and Median Burn Severity Over Years") +
  scale_color_manual(values = c("CMI Anomaly" = "blue", "Median Burn Severity" = "red")) +
  theme_minimal() +
  theme(legend.position = "bottom")

# Create all 1..5 lags relative to fire year t
lags <- 1:5

# If you ALREADY have df_cmi$fire_year (preferred):
if ("fire_year" %in% names(df_cmi)) {
  df_lags <- df_cmi %>%
    filter(fire_year >= min(df_sev$year), fire_year <= max(df_sev$year)) %>%
    mutate(lag = fire_year - water_year) %>%
    filter(lag %in% lags) %>%
    select(fire_year, lag, cmi_anom) %>%
    pivot_wider(names_from = lag, values_from = cmi_anom,
                names_prefix = "cmi_lag_") %>%
    arrange(fire_year) %>%
    rename(year = fire_year)
} else {
  # Map by arithmetic: for each fire year t, CMI anomaly at t-k
  df_lags <- expand_grid(year = df_sev$year, lag = lags) %>%
    mutate(water_year = year - lag) %>%
    left_join(df_cmi, by = "water_year") %>%
    select(year, lag, cmi_anom) %>%
    pivot_wider(names_from = lag, values_from = cmi_anom,
                names_prefix = "cmi_lag_") %>%
    arrange(year)
}


# Combine with severity
df_all <- df_sev %>%
  left_join(df_lags, by = "year")

# Keep only years with complete lag window
df_all <- df_all %>%
  filter(if_all(starts_with("cmi_lag_"), ~ !is.na(.x)))


# Equal weights across 1..5 years
w_equal <- rep(1/5, 5)

# Triangular weights (more weight to 4-5 years)
w_tri <- c(0.10, 0.15, 0.25, 0.25, 0.25)

# Use one set (switch if you prefer)
w <- w_equal  # or w_tri

lag_cols <- paste0("cmi_lag_", 1:5)

df_all <- df_all %>%
  mutate(
    cmi_star = as.numeric(as.matrix(select(., all_of(lag_cols))) %*% w)
  )


# Standardize for comparability (z-scores)
df_all <- df_all %>%
  mutate(
    sev_z = scale(mean_rbr)[,1],
    cmi_star_z = scale(cmi_star)[,1]
  )

# Convert to zoo
z_sev  <- zoo(df_all$sev_z, order.by = df_all$year)
z_cmi  <- zoo(df_all$cmi_star_z, order.by = df_all$year)

# LOESS smoothing helper for irregular yearly index
loess_smooth <- function(y, x, span = 0.4, degree = 1) {
  fit <- loess(y ~ x, span = span, degree = degree, surface = "direct", control = loess.control(trace.hat = "approx"))
  tibble(x = x, y = y, y_hat = predict(fit))
}

smooth_sev <- loess_smooth(y = coredata(z_sev), x = as.numeric(time(z_sev)), span = 0.4)
smooth_cmi <- loess_smooth(y = coredata(z_cmi), x = as.numeric(time(z_cmi)), span = 0.4)

# Merge smoothed curves back to calendar years
df_smooth <- tibble(
  year = index(z_sev),
  sev_z = coredata(z_sev),
  cmi_star_z = coredata(z_cmi),
  sev_loess = smooth_sev$y_hat,
  cmi_loess = smooth_cmi$y_hat
)

library(scales)
ggplot(df_smooth, aes(x = year)) +
  geom_line(aes(y = sev_z), color = "grey60", linewidth = 0.7, alpha = 0.6) +
  geom_line(aes(y = sev_loess), color = "#2c7fb8", linewidth = 1.4) +
  geom_line(aes(y = cmi_star_z), color = "grey70", linewidth = 0.7, alpha = 0.6) +
  geom_line(aes(y = cmi_loess), color = "#d95f0e", linewidth = 1.4) +
  scale_x_continuous(breaks = pretty_breaks()) +
  labs(x = "Year",
       y = "Standardized value (z-score)",
       title = "Median Burn Severity vs. Lag-weighted CMI Anomaly (LOESS-smoothed)",
       subtitle = "Both series standardized; LOESS span = 0.4") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(),
        plot.title.position = "plot")


# Gather long for easy iteration
df_long <- df_all %>%
  select(year, mean_rbr, all_of(lag_cols), cmi_star) %>%
  pivot_longer(cols = starts_with("cmi_lag_"),
               names_to = "lag_name", values_to = "cmi_anom_lag") %>%
  mutate(lag = as.integer(sub("cmi_lag_", "", lag_name)))

# Spearman correlation by lag
cor_by_lag <- df_long %>%
  group_by(lag) %>%
  summarize(
    n = n(),
    spearman_rho = cor(mean_rbr, cmi_anom_lag, method = "spearman", use = "complete.obs"),
    kendall_tau  = cor(mean_rbr, cmi_anom_lag, method = "kendall", use = "complete.obs"),
    .groups = "drop"
  ) %>%
  mutate(lag = factor(lag, levels = 1:5))

# With simple permutation CIs for Spearman (optional)
set.seed(123)
perm_spearman <- function(x, y, B = 5000) {
  obs <- suppressWarnings(cor(x, y, method = "spearman"))
  null <- replicate(B, {
    y_perm <- sample(y, replace = FALSE)
    suppressWarnings(cor(x, y_perm, method = "spearman"))
  })
  ci <- quantile(null, probs = c(0.025, 0.975), na.rm = TRUE)
  p  <- mean(abs(null) >= abs(obs))
  list(obs = obs, lwr = ci[1], upr = ci[2], p_perm = p)
}

cor_perm <- df_long %>%
  group_by(lag) %>%
  summarize({
    r <- perm_spearman(mean_rbr, cmi_anom_lag, B = 3000)
    tibble(spearman_obs = r$obs, ci_lwr = r$lwr, ci_upr = r$upr, p_perm = r$p_perm)
  }, .groups = "drop")

cor_by_lag$lag <- as.integer(as.character(cor_by_lag$lag))

cor_results <- left_join(cor_by_lag, cor_perm, by = "lag")

print(cor_results)

# Correlation with composite index (CMI*)
cor_composite <- tibble(
  n = nrow(df_all),
  spearman_rho = cor(df_all$mean_rbr, df_all$cmi_star, method = "spearman"),
  kendall_tau  = cor(df_all$mean_rbr, df_all$cmi_star, method = "kendall")
)
print(cor_composite)

 



ggplot(cor_results, aes(x = lag, y = spearman_obs)) +
  geom_hline(yintercept = 0, color = "grey70") +
  geom_point(size = 3, color = "#2c7fb8") +
  geom_errorbar(aes(ymin = ci_lwr, ymax = ci_upr), width = 0.15, color = "#2c7fb8") +
  labs(x = "Lag (years before fire)", y = "Spearman correlation",
       title = "Association between Median Severity and Lagged CMI Anomalies",
       subtitle = "Points: observed rho; bars: permutation null 95% CI") +
  theme_minimal(base_size = 12)




# read in rbr data
rbr <- readRDS(here("data", "cleaned", "rbr_cleaned.rds"))

rbr <- rbr |> 
    mutate(year = as.integer(gsub("_.*", "", GID)))

gam.rbr <- gam(median ~ s(year, k = 34), data = rbr, method = "REML")
summary(gam.rbr)

plot.zoo(rbr_ts)

# plot the gam fit
plot(gam.rbr, shade = TRUE, main = "GAM Fit for RBR Median over Years")

# correlation between continuous variables except vpd and L
```
```{r}
colnames(moisture_data)
cor_matrix <- moisture_data |> 
    select(c(2:21)) |> # remove first column and 15 to 24th column
    #select only numeric columns
    select(where(is.numeric)) |>
    cor()

# look at correlation between ndmi, evi and ndmi columns
cor_moist <- moisture_data |> 
    select(c(mean_ndmi, mean_evi, mean_ndvi)) |> drop_na() |> 
    cor()
# apply a heatmap to visualize the correlation matrix

cor_melt <- melt(cor_matrix)
ggplot(data = cor_melt, aes(x=Var1, y=Var2, fill=value)) +
    geom_tile() +
    scale_fill_viridis(option = "C", limits = c(-1, 1)) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
    labs(title = "Correlation Matrix Heatmap", x = "", y = "", fill = "Correlation")

# find instances where correlation is > 0.5 or < -0.5
high_cor <- cor_melt |> 
    filter(value > 0.5 | value < -0.5 & Var1 != Var2)
high_cor |> filter(Var1 == "pct_up" | Var2 == "pct_up")

```


```{r}
ggpairs(moisture_data |>  select(c(-1, -15:-24)) |> # remove first column and 15 to 24th column
    #select only numeric columns
    select(where(is.numeric)) )
```

# out of all variables with grep(90th), plot the relationship with median and pct_90 facet wrap by varable name

```{r}
moisture_data_long <- moisture_data |> 
    pivot_longer(cols = contains("90th"), names_to = "variable", values_to = "value")

ggplot(moisture_data_long, aes(x = value, y = median)) +
    geom_point(alpha = 0.5) +
    theme_minimal() +
    facet_wrap(~ variable, scales = "free_x") +
    labs(title = "Relationship with FWI duration", x = "Value", y = "Median") +
    scale_color_viridis(discrete = TRUE)

```


Get number of fires per year and make a histogram


```{r}
fires_per_year <- moisture_data |> 
    group_by(year) |> 
    summarise(num_fires = n())

print(fires_per_year, n = Inf)

ggplot(fires_per_year, aes(x = year, y = num_fires)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    theme_minimal() +
    labs(title = "Number of Fires per Year", x = "Year", y = "Number of Fires")
```

What is the fire size in year == 2022
```{r}
year_2022 <- moisture_data |> 
    filter(year == 2022)
summary(year_2022$fire_size)

```

how many fires per season

```{r}
seaosn_summary <- moisture_data |> 
    group_by(season) |> 
    summarise(num_fires = n())
```

Because it is heavily skewed to summer fires (1032 vs 56), we will just include summer fires (after july 1)


# fire size distribution 
```{r}
# how many fires with fire_size_ha > 200
mini_fires <- moisture_data |> 
    filter(fire_size_ha < 200)
nrow(mini_fires)

# small (as per stocks)
small_fires <- moisture_data |> 
    filter(fire_size_ha >= 200 & fire_size_ha < 500)
nrow(small_fires)

# medium (as per stocks)
medium_fires <- moisture_data |> 
    filter(fire_size_ha >= 500 & fire_size_ha < 10000)
nrow(medium_fires)

# large (as per stocks)
large_fires <- moisture_data |> 
    filter(fire_size_ha >= 10000)
nrow(large_fires)

```


```{r}
#plot median by mean evi using a loess smoother
ggplot(moisture_data, aes(x = mean_ndmi, y = pct_90)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "loess", color = "blue") +
    theme_minimal() +
    labs(title = "Median vs Mean EVI", x = "Mean EVI", y = "Median")
```


