---
title: "Exploratory Data Analysis and Descriptive Statistics"
author: "Jack A. Goldman"
date: "2025-12-26"
format: html
---


```{r}
library(ggplot2)
library(tidyr)
library(ggplot2)
library(viridis)
library(cowplot)
library(patchwork)
library(readr)
library(mgcv)
library(dlnm)
library(reshape2)
library(GGally)
library(dplyr)
library(here)
library(ggridges)
```


# Read in data

```{r}
moisture_data <- readRDS(here("data", "analysis_ready", "moisture_lagged_data.rds")) |> 
  filter(fire_size_ha > 200) |> 
  filter(ecoregion_lvl %in%c("LN", "BTL", "LSJ"))
```

# ecoregion levels
```{r}
unique(moisture_data$ecoregion)
# counts per ecoregion 
ecoregion_counts <- moisture_data |> 
    group_by(ecoregion_lvl) |> 
    summarise(count = n())

print(ecoregion_counts)
```

# fires per year

```{r}
fires_per_year <- moisture_data |> 
    group_by(fyear) |> 
    summarise(num_fires = n())
print(fires_per_year, n = Inf)

ggplot(fires_per_year, aes(x = fyear, y = num_fires)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    theme_minimal() +
    labs(title = "Number of Fires per Year", x = "Year", y = "Number of Fires")

# geom ridgline plot of median by year
m_ridge <- ggplot(moisture_data, aes(x = median, y = as.factor(fyear), fill = ..x..)) +
    geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
    scale_fill_viridis(name = "Median", option = "C") +
    theme_minimal() +
    labs(title = "Distribution of Median by Year", x = "Median", y = "Year")

#geom ridgline of pct_95 by year
q95_ridge <- ggplot(moisture_data, aes(x = pct_95, y = as.factor(fyear), fill = ..x..)) +
    geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
    scale_fill_viridis(name = "95th Percentile", option = "C") +
    theme_minimal() +
    labs(title = "Distribution of Burn Severiy Extremes by Year", x = "95th Percentile Burn Severity", y = "Year")

# create  table of fires per year using gt package

fy_tbl <- fires_per_year |> 
    rename(Year = fyear, `Number of Fires` = num_fires) |> 
    gt() |>
    tab_options(
        table.font.names = "Arial",
        table.font.size = px(8)
    ) 
           


# patchwork group ridgeline plts and table
library(patchwork)
p1 <- (m_ridge | q95_ridge) +
    plot_layout(heights = c(3, 1))



ggsave(here("figures", "fires_per_year_ridgeline.png"), width = 10, height = 6)

```

# variantion in severity across years and ecoregions

```{r}
year_eco <- moisture_data |> 
    select(c(year, ecoregion_lvl, median, pct_95, fire_size_ha, mean_evi)) |> 
    drop_na() |> 
    select(-c(mean_evi))

#  total hectares burned per year and ecoregion and total
burned_area_summary <- year_eco |> 
    group_by(year, ecoregion_lvl) |> 
    summarise(total_burned_ha = sum(fire_size_ha),
              mean_median = mean(median),
              mean_pct_95 = mean(pct_95),
              num_fires = n()) |> 
    ungroup()

# total burned area
sum(year_eco$fire_size_ha)

# total ecoregion
ecoregion_summary <- year_eco |> 
    group_by(ecoregion_lvl) |> 
    summarise(total_burned_ha = sum(fire_size_ha),
              mean_median = mean(median),
              mean_pct_95 = mean(pct_95),
              num_fires = n()) |> 
    ungroup()

```

#
```{r}

```

# range of median and pct_95 

```{r}
summary(moisture_data$median)
summary(moisture_data$pct_95)
range(moisture_data$median)
range(moisture_data$pct_95)
sd(moisture_data$median)
sd(moisture_data$pct_95)
```

# range of model response/predictors as well as sd as gt table
```{r}
vars <- moisture_data |> 
    select(cmi_wy, vpd, pct_conf, fwi_90th, mean_evi)

# back transform all vars (scaled mean and sd*2)


```


```{r}

cor(moisture_data$pct_up, moisture_data$cmi_wy, use="complete.obs")
cor(moisture_data$pct_up, moisture_data$vpd,    use="complete.obs")


summary(moisture_df$pct_up); summary(moisture_df$pct_up)
range(moisture_df$pct_up);  range(moisture_df$pct_up)


```

# correlation between continuous variables except vpd and L

```{r}
colnames(moisture_data)
cor_matrix <- moisture_data |> 
    select(c(2:21)) |> # remove first column and 15 to 24th column
    #select only numeric columns
    select(where(is.numeric)) |>
    cor()

# look at correlation between ndmi, evi and ndmi columns
cor_moist <- moisture_data |> 
    select(c(mean_ndmi, mean_evi, mean_ndvi)) |> drop_na() |> 
    cor()
# apply a heatmap to visualize the correlation matrix

cor_melt <- melt(cor_matrix)
ggplot(data = cor_melt, aes(x=Var1, y=Var2, fill=value)) +
    geom_tile() +
    scale_fill_viridis(option = "C", limits = c(-1, 1)) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
    labs(title = "Correlation Matrix Heatmap", x = "", y = "", fill = "Correlation")

# find instances where correlation is > 0.5 or < -0.5
high_cor <- cor_melt |> 
    filter(value > 0.5 | value < -0.5 & Var1 != Var2)
high_cor |> filter(Var1 == "pct_up" | Var2 == "pct_up")

```


```{r}
ggpairs(moisture_data |>  select(c(-1, -15:-24)) |> # remove first column and 15 to 24th column
    #select only numeric columns
    select(where(is.numeric)) )
```

# out of all variables with grep(90th), plot the relationship with median and pct_90 facet wrap by varable name

```{r}
moisture_data_long <- moisture_data |> 
    pivot_longer(cols = contains("90th"), names_to = "variable", values_to = "value")

ggplot(moisture_data_long, aes(x = value, y = median)) +
    geom_point(alpha = 0.5) +
    theme_minimal() +
    facet_wrap(~ variable, scales = "free_x") +
    labs(title = "Relationship with FWI duration", x = "Value", y = "Median") +
    scale_color_viridis(discrete = TRUE)

```


Get number of fires per year and make a histogram


```{r}
fires_per_year <- moisture_data |> 
    group_by(year) |> 
    summarise(num_fires = n())

print(fires_per_year, n = Inf)

ggplot(fires_per_year, aes(x = year, y = num_fires)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    theme_minimal() +
    labs(title = "Number of Fires per Year", x = "Year", y = "Number of Fires")
```

What is the fire size in year == 2022
```{r}
year_2022 <- moisture_data |> 
    filter(year == 2022)
summary(year_2022$fire_size)

```

how many fires per season

```{r}
seaosn_summary <- moisture_data |> 
    group_by(season) |> 
    summarise(num_fires = n())
```

Because it is heavily skewed to summer fires (1032 vs 56), we will just include summer fires (after july 1)


# fire size distribution 
```{r}
# how many fires with fire_size_ha > 200
mini_fires <- moisture_data |> 
    filter(fire_size_ha < 200)
nrow(mini_fires)

# small (as per stocks)
small_fires <- moisture_data |> 
    filter(fire_size_ha >= 200 & fire_size_ha < 500)
nrow(small_fires)

# medium (as per stocks)
medium_fires <- moisture_data |> 
    filter(fire_size_ha >= 500 & fire_size_ha < 10000)
nrow(medium_fires)

# large (as per stocks)
large_fires <- moisture_data |> 
    filter(fire_size_ha >= 10000)
nrow(large_fires)

```


```{r}
#plot median by mean evi using a loess smoother
ggplot(moisture_data, aes(x = mean_ndmi, y = pct_90)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "loess", color = "blue") +
    theme_minimal() +
    labs(title = "Median vs Mean EVI", x = "Mean EVI", y = "Median")
```


```{r}
# how many fires inside vs outside aou
aou_summary <- moisture_data |> 
    group_by(aou) |> 
    summarise(num_fires = n())
print(aou_summary)
```