```{r}
library(ggplot2)
library(patchwork)
# source
source(here("utils", "plot_functions.r"))
```


# Linear Predictor plots 





# H4: are there lagged effects?

Analyzed lagged effects through visualization of the 2D smooths from the distributed lag smooths. Significant areas circled in black . If no significant areas, there were no differences across lags. Suggests that lagged effects are not as importance as the overall effects of moisture condditions on fire severity.

## set colour palettes for ecoregions

```{r}
lsj <- scale_fill_viridis_c(
        name = 'p(Median)',
        option = "D",
        guide = guide_colorbar(
          direction = "horizontal",
          title.position = "top",
          title.hjust = 0.5
        )
      ) 
ln <- scale_fill_viridis_c(
        name = 'p(Median)',
        option = "E",
        guide = guide_colorbar(
          direction = "horizontal",
          title.position = "top",
          title.hjust = 0.5
        )
      ) 
lsj_e <- scale_fill_viridis_c(
        name = 'p(Extreme)',
        option = "D",
        guide = guide_colorbar(
          direction = "horizontal",
          title.position = "top",
          title.hjust = 0.5
        )
      ) 
ln_e<- scale_fill_viridis_c(
        name = 'p(Extreme)',
        option = "E",
        guide = guide_colorbar(
          direction = "horizontal",
          title.position = "top",
          title.hjust = 0.5
        )
      ) 
```

## Median
```{r}


estimates_btl <- get_lag_estimates(m_vpd_median, lag_var = "L", smooth_var = "vpd", weights = "weights_btl")
plot_vpd_btl <- plot_lag_smooth(m_vpd_median, estimates_btl, "vpd", "L", "weights_btl")
plot_vpd_btl

estimates_lsj <- get_lag_estimates(m_vpd_median, lag_var = "L", smooth_var = "vpd", weights = "weights_lsj")
plot_vpd_lsj <- plot_lag_smooth(m_vpd_median, estimates_lsj, "vpd", "L", "weights_lsj")
plot_vpd_lsj + lsj

estimates_ln <- get_lag_estimates(m_vpd_median, lag_var = "L", smooth_var = "vpd", weights = "weights_ln")
plot_vpd_ln <- plot_lag_smooth(m_vpd_median, estimates_ln, "vpd", "L", "weights_ln")
plot_vpd_ln + ln


```

## cmi
```{r}
estimates_cmi <- get_lag_estimates(m_cmi_median,  "LL",  "cmi_wy", "weights_btl2")
plot_cmi <- plot_lag_smooth(m_cmi_median, estimates_cmi, "cmi_wy", "LL", "weights_btl2")

plot_cmi

estimates_cmi_lsj <- get_lag_estimates(m_cmi_median, lag_var = "LL", smooth_var = "cmi_wy", weights = "weights_lsj2")
plot_cmi_lsj <- plot_lag_smooth(m_cmi_median, estimates_cmi_lsj, "cmi_wy", "LL", "weights_lsj2")

plot_cmi_lsj + lsj

estimates_cmi_ln <- get_lag_estimates(m_cmi_median, lag_var = "LL", smooth_var = "cmi_wy", weights = "weights_ln2")
plot_cmi_ln <- plot_lag_smooth(m_cmi_median, estimates_cmi_ln, "cmi_wy", "LL", "weights_ln2")

plot_cmi_ln + ln

```

# Extreme


```{r}
estimates_btl_ext <- get_lag_estimates(m_vpd_ext, lag_var = "L", smooth_var = "vpd", weights = "weights_btl")
plot_vpd_btl_ext <- plot_lag_smooth(m_vpd_ext, estimates_btl_ext, "vpd", "L", "weights_btl")
plot_vpd_btl_ext

estimates_lsj_ext <- get_lag_estimates(m_vpd_ext, lag_var = "L", smooth_var = "vpd", weights = "weights_lsj")
plot_vpd_lsj_ext <- plot_lag_smooth(m_vpd_ext, estimates_lsj_ext, "vpd", "L", "weights_lsj")
plot_vpd_lsj_ext + lsj_e

estimates_ln_ext <- get_lag_estimates(m_vpd_ext, lag_var = "L", smooth_var = "vpd", weights = "weights_ln")
plot_vpd_ln_ext <- plot_lag_smooth(m_vpd_ext, estimates_ln_ext, "vpd", "L", "weights_ln")
plot_vpd_ln_ext + ln_e
```


```{r}
estimates_cmi_ext <- get_lag_estimates(m_cmi_ext,  "LL",  "cmi_wy", "weights_btl2")
plot_cmi_ext <- plot_lag_smooth(m_cmi_ext, estimates_cmi_ext, "cmi_wy", "LL", "weights_btl2") 
plot_cmi_ext

estimates_cmi_lsj_ext <- get_lag_estimates(m_cmi_ext, lag_var = "LL", smooth_var = "cmi_wy", weights = "weights_lsj2")  
plot_cmi_lsj_ext <- plot_lag_smooth(m_cmi_ext, estimates_cmi_lsj_ext, "cmi_wy", "LL", "weights_lsj2")
plot_cmi_lsj_ext + lsj_e

estimates_cmi_ln_ext <- get_lag_estimates(m_cmi_ext, lag_var = "LL", smooth_var = "cmi_wy", weights = "weights_ln2")  
plot_cmi_ln_ext <- plot_lag_smooth(m_cmi_ext, estimates_cmi_ln_ext, "cmi_wy", "LL", "weights_ln2")
plot_cmi_ln_ext + ln_e
```


# Cumulative effects plots


