```{r}
library(ggplot2)
library(patchwork)
library(gratia)
# source
source(here("utils", "plot_functions.r"))
```


## Fuels plots

We plotted the linear predictor from the CMI model because it had slightly better fit for both median and extremes

## mean EVI

```{r}


# For median burn severity (m_cmi_median)
# Get estimates for each ecoregion's pct_conf smooth
smooth_btl <- smooth_estimates(m_cmi_median, smooth = "s(mean_evi):ecoregion_lvlBTL", n = 100, partial = TRUE, unconditional = TRUE) |> add_confint()
smooth_ln <- smooth_estimates(m_cmi_median, smooth = "s(mean_evi):ecoregion_lvlLN", n = 100, partial = TRUE, unconditional = TRUE)  |> add_confint()
smooth_lsj <- smooth_estimates(m_cmi_median, smooth = "s(mean_evi):ecoregion_lvlLSJ", n = 100, partial = TRUE, unconditional = TRUE) |> add_confint()

# Combine into one data frame for plotting
smooth_med <- bind_rows(
  smooth_btl %>% mutate(ecoregion = "BTL"),
  smooth_ln %>% mutate(ecoregion = "LN"),
  smooth_lsj %>% mutate(ecoregion = "LSJ"))

# Plot with overlaid lines per ecoregion
evi_med <- ggplot() +
  #geom_point(data = moisture_data, aes(x = mean_evi, y = median, color = ecoregion_lvl), alpha = 0.1) +
  geom_line(data = smooth_med, aes(x = mean_evi, y = .estimate, color = ecoregion), size = 1.5, lineend = "round") +
  geom_ribbon(data = smooth_med, aes(x = mean_evi, ymin = .lower_ci, ymax = .upper_ci, fill = ecoregion), alpha = 0.2) +
  labs(x = "Mean EVI (z-score)", title = "VPD", y = "Median Burn Severity", color = "Ecoregion", fill = "Ecoregion") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_color_manual(values = c("BTL" = "gray60", "LN" = "darkorange", "LSJ" = "forestgreen")) +
  scale_fill_manual(values = c("BTL" = "gray60", "LN" = "darkorange", "LSJ" = "forestgreen"))



# For extreme burn severity (m_cmi_ext)
smooth_btl_ext <- smooth_estimates(m_cmi_ext, smooth = "s(mean_evi):ecoregion_lvlBTL", n = 100, partial = TRUE, unconditional = TRUE) |> add_confint()
smooth_ln_ext <- smooth_estimates(m_cmi_ext, smooth = "s(mean_evi):ecoregion_lvlLN", n = 100, partial = TRUE, unconditional = TRUE) |> add_confint()
smooth_lsj_ext <- smooth_estimates(m_cmi_ext, smooth = "s(mean_evi):ecoregion_lvlLSJ", n = 100, partial = TRUE, unconditional = TRUE) |> add_confint()

smooth_ext <- bind_rows(
  smooth_btl_ext %>% mutate(ecoregion = "BTL"),
  smooth_ln_ext %>% mutate(ecoregion = "LN"),
  smooth_lsj_ext %>% mutate(ecoregion = "LSJ")
)

evi_ext <- ggplot() +
  #geom_point(data = moisture_data, aes(x = mean_evi, y = pct_95, color = ecoregion_lvl), alpha = 0.1) +
  geom_line(data = smooth_ext, aes(x = mean_evi, y = .estimate, color = ecoregion), size = 1.5, lineend = "round") +
  geom_ribbon(data = smooth_ext, aes(x = mean_evi, ymin = .lower_ci, ymax = .upper_ci, fill = ecoregion), alpha = 0.2) +
  labs(x = "Mean EVI (z-score)", title = "CMI", y = "Extreme Burn Severity", color = "Ecoregion", fill = "Ecoregion") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_color_manual(values = c("BTL" = "gray60", "LN" = "darkorange", "LSJ" = "forestgreen")) +
  scale_fill_manual(values = c("BTL" = "gray60", "LN" = "darkorange", "LSJ" = "forestgreen"))

evi_ext

```


```{r}

# For median burn severity (m_cmi_median)
# Get estimates for each ecoregion's pct_conf smooth
smooth_btl <- smooth_estimates(m_cmi_median, smooth = "s(pct_conf):ecoregion_lvlBTL", n = 100, partial = TRUE, unconditional = TRUE) |> add_confint()
smooth_ln <- smooth_estimates(m_cmi_median, smooth = "s(pct_conf):ecoregion_lvlLN", n = 100, partial = TRUE, unconditional = TRUE)  |> add_confint()
smooth_lsj <- smooth_estimates(m_cmi_median, smooth = "s(pct_conf):ecoregion_lvlLSJ", n = 100, partial = TRUE, unconditional = TRUE) |> add_confint()

# Combine into one data frame for plotting
smooth_med <- bind_rows(
  smooth_btl %>% mutate(ecoregion = "BTL"),
  smooth_ln %>% mutate(ecoregion = "LN"),
  smooth_lsj %>% mutate(ecoregion = "LSJ"))

# Plot with overlaid lines per ecoregion
conf_med <- ggplot() +
  #geom_point(data = moisture_data, aes(x = pct_conf, y = median, color = ecoregion_lvl), alpha = 0.1) +
  geom_line(data = smooth_med, aes(x = pct_conf, y = .estimate, color = ecoregion), size = 1.5, lineend = "round") +
  geom_ribbon(data = smooth_med, aes(x = pct_conf, ymin = .lower_ci, ymax = .upper_ci, fill = ecoregion), alpha = 0.2) +
  labs(x = "Percent Conifer Cover (z-score)", title = "VPD" ,y = "Median Burn Severity", color = "Ecoregion", fill = "Ecoregion") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_color_manual(values = c("BTL" = "gray60", "LN" = "darkorange", "LSJ" = "forestgreen")) +
  scale_fill_manual(values = c("BTL" = "gray60", "LN" = "darkorange", "LSJ" = "forestgreen"))



# For extreme burn severity (m_cmi_ext)
smooth_btl_ext <- smooth_estimates(m_cmi_ext, smooth = "s(pct_conf):ecoregion_lvlBTL", n = 100, partial = TRUE, unconditional = TRUE) |> add_confint()
smooth_ln_ext <- smooth_estimates(m_cmi_ext, smooth = "s(pct_conf):ecoregion_lvlLN", n = 100, partial = TRUE, unconditional = TRUE) |> add_confint()
smooth_lsj_ext <- smooth_estimates(m_cmi_ext, smooth = "s(pct_conf):ecoregion_lvlLSJ", n = 100, partial = TRUE, unconditional = TRUE) |> add_confint()

smooth_ext <- bind_rows(
  smooth_btl_ext %>% mutate(ecoregion = "BTL"),
  smooth_ln_ext %>% mutate(ecoregion = "LN"),
  smooth_lsj_ext %>% mutate(ecoregion = "LSJ")
)

conf_ext <- ggplot() +
  #geom_point(data = moisture_data, aes(x = pct_conf, y = pct_95, color = ecoregion_lvl), alpha = 0.1) +
  geom_line(data = smooth_ext, aes(x = pct_conf, y = .estimate, color = ecoregion), size = 1.5, lineend = "round") +
  geom_ribbon(data = smooth_ext, aes(x = pct_conf, ymin = .lower_ci, ymax = .upper_ci, fill = ecoregion), alpha = 0.2) +
  labs(x = "Percent Conifer Cover (z-score)", title = "CMI", y = "Extreme Burn Severity", color = "Ecoregion", fill = "Ecoregion") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_color_manual(values = c("BTL" = "gray60", "LN" = "darkorange", "LSJ" = "forestgreen")) +
  scale_fill_manual(values = c("BTL" = "gray60", "LN" = "darkorange", "LSJ" = "forestgreen"))

conf_ext

```
```{r}
conf_med + conf_ext + evi_med + evi_ext + plot_layout(ncol = 2) + 
  plot_annotation(tag_levels = "A")

# save
ggsave(here("figures", "linear_predictor_plots_conifer_evi_smooth.png"), width = 10, height = 8)
```


# Deviance explained by each group

```{r}
as.data.frame(cmi_ext_hp$hierarchical.partitioning)

I.cmi.ext <- cmi_ext_hp$hierarchical.partitioning |> 
    as.data.frame() |> 
    rownames_to_column(var = "Variable") |> mutate(Variable = 
    case_when(
        Variable == "fire_weather" ~ "Fire Weather", 
        Variable == "fuels" ~ "Fuels",
        Variable == "moisture" ~ "Moisture"))

I.cmi.median <- cmi_median_hp$hierarchical.partitioning |> 
    as.data.frame() |> 
    rownames_to_column(var = "Variable") |> mutate(Variable = 
    case_when(
        Variable == "fire_weather" ~ "Fire Weather", 
        Variable == "fuels" ~ "Fuels",
        Variable == "moisture" ~ "Moisture"))
I.vpd.ext <- vpd_ext_hp$hierarchical.partitioning |> 
    as.data.frame() |> 
    rownames_to_column(var = "Variable") |> mutate(Variable = 
    case_when(
        Variable == "fire_weather" ~ "Fire Weather", 
        Variable == "fuels" ~ "Fuels",
        Variable == "moisture" ~ "Moisture"))
I.vpd.median <- vpd_median_hp$hierarchical.partitioning |> 
    as.data.frame() |> 
    rownames_to_column(var = "Variable") |> mutate(Variable = 
    case_when(
        Variable == "fire_weather" ~ "Fire Weather", 
        Variable == "fuels" ~ "Fuels",
        Variable == "moisture" ~ "Moisture"))


# make 4 individual bar plots of I.perc(%) for each model where the x axis is I.perc(%) and fill is Variable

p.cmi.med <- ggplot(I.cmi.median, aes(x = Variable, y = `I.perc(%)`, fill = Variable)) +
  geom_col() +
  geom_text(aes(label = paste0(round(`I.perc(%)`, 1), "%")), vjust = -0.5, size = 3.5) +
  labs(title = "CMI Median", y = "Deviance Explained (%)", x = "") +
  theme_bw() +
  coord_cartesian(ylim = c(0, 100)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        plot.title = element_text(hjust = 0.5, size = 12, face = "bold")) +
  scale_fill_manual(values = c("Moisture" = "gray62", "Fire Weather" = "#CD0BBC", "Fuels" = "#F5C710")) +
  guides(fill = "none")

p.cmi.ext <- ggplot(I.cmi.ext, aes(x = Variable, y = `I.perc(%)`, fill = Variable)) +
  geom_col() +
  geom_text(aes(label = paste0(round(`I.perc(%)`, 1), "%")), vjust = -0.5, size = 3.5) +
  labs(title = "CMI Extreme", y = NULL,  x = "") +
  theme_bw() +
  coord_cartesian(ylim = c(0, 100)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        plot.title = element_text(hjust = 0.5, size = 12, face = "bold")) +
  scale_fill_manual(values = c("Moisture" = "gray62", "Fire Weather" = "#CD0BBC", "Fuels" = "#F5C710")) +
  guides(fill = "none")

p.vpd.med <- ggplot(I.vpd.median, aes(x = Variable, y = `I.perc(%)`, fill = Variable)) +
  geom_col() +
  geom_text(aes(label = paste0(round(`I.perc(%)`, 1), "%")), vjust = -0.5, size = 3.5) +
  labs(title = "VPD Median", y = "Deviance Explained (%)", x = "") +
  theme_bw() +
  coord_cartesian(ylim = c(0, 100)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        plot.title = element_text(hjust = 0.5, size = 12, face = "bold")) +
  scale_fill_manual(values = c("Moisture" = "gray62", "Fire Weather" = "#CD0BBC", "Fuels" = "#F5C710")) +
  guides(fill = "none")

p.vpd.ext <- ggplot(I.vpd.ext, aes(x = Variable, y = `I.perc(%)`, fill = Variable)) +
  geom_col() +
  geom_text(aes(label = paste0(round(`I.perc(%)`, 1), "%")), vjust = -0.5, size = 3.5) +
  labs(title = "VPD Extreme", y = NULL, x = "") +
  theme_bw() +
  coord_cartesian(ylim = c(0, 100)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        plot.title = element_text(hjust = 0.5, size = 12, face = "bold")) +
  scale_fill_manual(values = c("Moisture" = "gray62", "Fire Weather" = "#CD0BBC", "Fuels" = "#F5C710")) +
  guides(fill = "none")

p.cmi.med <- place_label_inside(p.cmi.med, "a)")
p.cmi.ext <- place_label_inside(p.cmi.ext, "b)")
p.vpd.med <- place_label_inside(p.vpd.med, "c)")
p.vpd.ext <- place_label_inside(p.vpd.ext, "d)")

# combine plots using patchwork
 p.vpd.med + p.vpd.ext + p.cmi.med + p.cmi.ext + plot_layout(ncol = 2) + 
  plot_annotation(tag_levels = "A")

# save
ggsave(here("figures", "deviance_explained_by_variable_group_v2.png"), width = 5, height = 5)


```

# H4: are there lagged effects?

Analyzed lagged effects through visualization of the 2D smooths from the distributed lag smooths. Significant areas circled in black . If no significant areas, there were no differences across lags. Suggests that lagged effects are not as importance as the overall effects of moisture condditions on fire severity.

## set colour palettes for ecoregions

```{r}
lsj <- scale_fill_viridis_c(
        name = 'Median\nBurn Severity',
        option = "D",
        guide = guide_colorbar(
          direction = "vertical",
          title.position = "top",
          title.hjust = 0.5
        )
      ) 
ln <- scale_fill_viridis_c(
        name = 'Median\nBurn Severity',
        option = "F",
        guide = guide_colorbar(
          direction = "vertical",
          title.position = "top",
          title.hjust = 0.5
        )
      ) 
lsj_e <- scale_fill_viridis_c(
        name = 'Extreme\nBurn Severity',
        option = "D",
        guide = guide_colorbar(
          direction = "vertical",
          title.position = "top",
          title.hjust = 0.5
        )
      ) 
ln_e<- scale_fill_viridis_c(
        name = 'Extreme\nBurn Severity',
        option = "F",
        guide = guide_colorbar(
          direction = "vertical",
          title.position = "top",
          title.hjust = 0.5
        )
      ) 
```

## Median
```{r}


estimates_btl <- get_lag_estimates(m_vpd_median, lag_var = "L", smooth_var = "vpd", weights = "weights_btl")
plot_vpd_btl <- plot_lag_smooth(m_vpd_median, estimates_btl, "vpd", "L", "weights_btl")
plot_vpd_btl

estimates_lsj <- get_lag_estimates(m_vpd_median, lag_var = "L", smooth_var = "vpd", weights = "weights_lsj")
plot_vpd_lsj <- plot_lag_smooth(m_vpd_median, estimates_lsj, "vpd", "L", "weights_lsj")
plot_vpd_lsj <- plot_vpd_lsj + lsj

estimates_ln <- get_lag_estimates(m_vpd_median, lag_var = "L", smooth_var = "vpd", weights = "weights_ln")
plot_vpd_ln <- plot_lag_smooth(m_vpd_median, estimates_ln, "vpd", "L", "weights_ln")
plot_vpd_ln <- plot_vpd_ln + ln


```

## cmi
```{r}
estimates_cmi <- get_lag_estimates(m_cmi_median,  "LL",  "cmi_wy", "weights_btl2")
plot_cmi <- plot_lag_smooth(m_cmi_median, estimates_cmi, "cmi_wy", "LL", "weights_btl2")

plot_cmi

estimates_cmi_lsj <- get_lag_estimates(m_cmi_median, lag_var = "LL", smooth_var = "cmi_wy", weights = "weights_lsj2")
plot_cmi_lsj <- plot_lag_smooth(m_cmi_median, estimates_cmi_lsj, "cmi_wy", "LL", "weights_lsj2")

plot_cmi_lsg <- plot_cmi_lsj + lsj

estimates_cmi_ln <- get_lag_estimates(m_cmi_median, lag_var = "LL", smooth_var = "cmi_wy", weights = "weights_ln2")
plot_cmi_ln <- plot_lag_smooth(m_cmi_median, estimates_cmi_ln, "cmi_wy", "LL", "weights_ln2")

plot_cmi_ln <- plot_cmi_ln + ln

```

# Extreme


```{r}
estimates_btl_ext <- get_lag_estimates(m_vpd_ext, lag_var = "L", smooth_var = "vpd", weights = "weights_btl")
plot_vpd_btl_ext <- plot_lag_smooth(m_vpd_ext, estimates_btl_ext, "vpd", "L", "weights_btl")
plot_vpd_btl_ext

estimates_lsj_ext <- get_lag_estimates(m_vpd_ext, lag_var = "L", smooth_var = "vpd", weights = "weights_lsj")
plot_vpd_lsj_ext <- plot_lag_smooth(m_vpd_ext, estimates_lsj_ext, "vpd", "L", "weights_lsj")
plot_vpd_lsj_ext <- plot_vpd_lsj_ext + lsj_e

estimates_ln_ext <- get_lag_estimates(m_vpd_ext, lag_var = "L", smooth_var = "vpd", weights = "weights_ln")
plot_vpd_ln_ext <- plot_lag_smooth(m_vpd_ext, estimates_ln_ext, "vpd", "L", "weights_ln")
plot_vpd_ln_ext <- plot_vpd_ln_ext + ln_e
```


```{r}
estimates_cmi_ext <- get_lag_estimates(m_cmi_ext,  "LL",  "cmi_wy", "weights_btl2")
plot_cmi_ext <- plot_lag_smooth(m_cmi_ext, estimates_cmi_ext, "cmi_wy", "LL", "weights_btl2") 
plot_cmi_ext

estimates_cmi_lsj_ext <- get_lag_estimates(m_cmi_ext, lag_var = "LL", smooth_var = "cmi_wy", weights = "weights_lsj2")  
plot_cmi_lsj_ext <- plot_lag_smooth(m_cmi_ext, estimates_cmi_lsj_ext, "cmi_wy", "LL", "weights_lsj2")
plot_cmi_lsj_ext <- plot_cmi_lsj_ext + lsj_e

estimates_cmi_ln_ext <- get_lag_estimates(m_cmi_ext, lag_var = "LL", smooth_var = "cmi_wy", weights = "weights_ln2")  
plot_cmi_ln_ext <- plot_lag_smooth(m_cmi_ext, estimates_cmi_ln_ext, "cmi_wy", "LL", "weights_ln2")
plot_cmi_ln_ext <- plot_cmi_ln_ext + ln_e 
```

patchwork them together by median and ext
```{r}

# median

top1 <- plot_vpd_btl + ggtitle("BTL") + theme(plot.title = element_text(face = "bold", hjust = 0.5))
top2 <- plot_vpd_lsj + ggtitle("LSJ") + theme(plot.title = element_text(face = "bold", hjust = 0.5))
top3 <- plot_vpd_ln  + ggtitle("LN")  + theme(plot.title = element_text(face = "bold", hjust = 0.5))

p <- top1 + top2 + top3 + plot_cmi + plot_cmi_lsg + plot_cmi_ln +
    plot_layout(ncol = 3) + plot_annotation(tag_levels = "A")

p

#save to figures
ggsave(here("figures", "lagged_effects_median_v2.png"), width = 12, height = 8, dpi = 300)

# extreme
top1_e <- plot_vpd_btl_ext + ggtitle("BTL") + theme(plot.title = element_text(face = "bold", hjust = 0.5))
top2_e <- plot_vpd_lsj_ext + ggtitle("LSJ") + theme(plot.title = element_text(face = "bold", hjust = 0.5))
top3_e <- plot_vpd_ln_ext  + ggtitle("LN")  + theme(plot.title = element_text(face = "bold", hjust = 0.5))

p_e <- top1_e + top2_e + top3_e + plot_cmi_ext + plot_cmi_lsj_ext + plot_cmi_ln_ext +
    plot_layout(ncol = 3) + plot_annotation(tag_levels = "A")

p_e
#save to figures
ggsave(here("figures", "lagged_effects_extreme_v2.png"), width = 12, height = 8, dpi = 300)

```

# steep slops

## BTL
```{r}
# median vs extreme - vpd
estimates_btl_steep <- get_steepest(estimates_btl)
estimates_btl_ext <-  get_steepest(estimates_btl_ext)

estimates_cmi_steep <-  estimates_cmi |> filter(`.estimate` == min(`.estimate`))
estimates_cmi_ext_steep <- estimates_cmi_ext |> filter(tf == FALSE) |> filter(`.estimate` == min(`.estimate`)) 

```

## Lake Nipigon
```{r}
# Lake Nipigon
# median vs extreme - ln vpd
ln_steep_med <-  get_steepest(estimates_ln) 
ln_steep_ext <-  get_steepest(estimates_ln_ext)

# median vs extreme - ln cmi
cmi_ln_steep_med <-  get_steepest(estimates_cmi_ln) 
cmi_ln_steep_ext <-  get_steepest(estimates_cmi_ln_ext) 

```

## LSJ
```{r}
#  extreme lsj -cmi
estimates_lsj_steep <-  estimates_lsj_ext |> filter(tf == FALSE)|> filter(`.estimate` == max(`.estimate`))
estimates_cmi_lsj_steep <-  estimates_cmi_lsj_ext |> filter(`.estimate` == min(`.estimate`))


```


```{r}
library(gratia)
library(ggplot2)
library(dplyr)

# Load your model (e.g., m_vpd_median) and data
# Assume moisture_data is loaded

# Define low and high pct_conf levels (e.g., based on quantiles)
low_pct_conf <- quantile(moisture_data$pct_conf, 0.25, na.rm = TRUE)  # e.g., 25th percentile for "low"
high_pct_conf <- quantile(moisture_data$pct_conf, 0.75, na.rm = TRUE)  # 75th percentile for "high"

# Slice data for lagged VPD effects at fixed pct_conf (set other covariates to means)
ds_low <- data_slice(m_vpd_median, 
                     vpd = evenly(vpd, n = 50), 
                     L = evenly(L, n = 10), 
                     pct_conf = low_pct_conf,  # Fix at low conifer
                     mean_evi = mean(moisture_data$mean_evi, na.rm = TRUE),
                     fwi_90th = mean(moisture_data$fwi_90th, na.rm = TRUE),
                     weights_btl = 1)  # Assuming BTL for simplicity

ds_high <- data_slice(m_vpd_median, 
                      vpd = evenly(vpd, n = 50), 
                      L = evenly(L, n = 10), 
                      pct_conf = high_pct_conf,  # Fix at high conifer
                      mean_evi = mean(moisture_data$mean_evi, na.rm = TRUE),
                      fwi_90th = mean(moisture_data$fwi_90th, na.rm = TRUE),
                      weights_btl = 1)

# Get fitted values
fv_low <- fitted_values(m_vpd_median, data = ds_low, scale = "response")
fv_high <- fitted_values(m_vpd_median, data = ds_high, scale = "response")

# Plot: Compare predicted severity vs. VPD at low vs. high pct_conf
ggplot() +
  geom_line(data = fv_low, aes(x = vpd, y = .fitted, color = "Low Conifer"), size = 1) +
  geom_line(data = fv_high, aes(x = vpd, y = .fitted, color = "High Conifer"), size = 1) +
  labs(x = "VPD (z-score)", y = "Predicted Median Burn Severity", 
       title = "Moisture Effect Conditional on Conifer Cover") +
  scale_color_manual(values = c("Low Conifer" = "red", "High Conifer" = "blue")) +
  theme_minimal()
deriv_low <- derivatives(m_vpd_median, data = ds_low, type = "central")

model.data <- as.data.frame(m_vpd_median$model)

ds_low <- data_slice(m_vpd_median, 
                     vpd = evenly(vpd, n = 50), 
                     L = evenly(L, n = 10), 
                     pct_conf = low_pct_conf,
                     mean_evi = mean(moisture_data$mean_evi, na.rm = TRUE),
                     fwi_90th = mean(moisture_data$fwi_90th, na.rm = TRUE),
                     weights_btl = 1,
                     weights_ln = 0,  # Include all weight vars
                     weights_lsj = 0,
                     fyear = factor(levels(model.data$fyear)[1]),  # Add missing vars
                     ecoregion_lvl = factor(levels(model.data$ecoregion_lvl)[1]))

ds_high <- data_slice(m_vpd_median, 
                      vpd = evenly(vpd, n = 50), 
                      L = evenly(L, n = 10), 
                      pct_conf = high_pct_conf,
                      mean_evi = mean(moisture_data$mean_evi, na.rm = TRUE),
                      fwi_90th = mean(moisture_data$fwi_90th, na.rm = TRUE),
                      weights_btl = 1,
                      weights_ln = 0,
                      weights_lsj = 0,
                      fyear = factor(levels(model.data$fyear)[1]),
                      ecoregion_lvl = factor(levels(model.data$ecoregion_lvl)[1]))

# Now derivatives should work
deriv_low <- derivatives(m_vpd_median, select = "te(vpd,L):weights_btl", data = ds_low, type = "central")
deriv_high <- derivatives(m_vpd_median, data = ds_high, type = "central")
```



# Cumulative effects plots



```{r}
#Assuming estimates_btl is already loaded
avg_effect <- estimates_btl |> 
  filter(tf == FALSE & L == 16) |> 
  group_by(vpd) |> 
  summarise(
    avg_est = mean(.estimate, na.rm = TRUE),
    se_mean = sd(.estimate, na.rm = TRUE) / sqrt(n()),
    lower_ci = avg_est - 1.96 * se_mean,
    upper_ci = avg_est + 1.96 * se_mean
  )

ggplot(avg_effect, aes(x = vpd, y = avg_est)) +
  geom_line(color = "blue") +
  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci), alpha = 0.2, fill = "blue") +
  labs(title = "Average Effect of VPD Across All Lags (BTL)", 
       x = "VPD (z-score)", 
       y = "Average Estimated Effect on Severity") +
  theme_minimal()


draw(cv)

min_median <- min(moisture_data$median, na.rm = TRUE)
max_median <- max(moisture_data$median, na.rm = TRUE)
min_vpd <- min(moisture_data$vpd, na.rm = TRUE)
max_vpd <- max(moisture_data$vpd, na.rm = TRUE)

model.data <- as.data.frame(m_vpd_median$model)

# Create the prediction dataframe
pred_data <- expand.grid(
  vpd = seq(min_vpd, max_vpd, length.out = 100),
  L = seq(1, 30, by = 1),
  weights_btl = 1,
  weights_ln = 0,
  weights_lsj = 0,
  pct_conf = mean(model.data$pct_conf, na.rm = TRUE),
  fwi_90th = mean(model.data$fwi_90th, na.rm = TRUE),
  mean_evi = mean(model.data$mean_evi, na.rm = TRUE),
  fyear = factor(levels(model.data$fyear)[1]),
  ecoregion_lvl = factor(levels(model.data$ecoregion_lvl)[1])
)
ds <- data_slice(m_cmi_median, cmi_wy = evenly(cmi_wy, n = 100), LL = evenly(LL, n = 30))

fv1 <- fitted_values(m_cmi_median, data = ds, scale = "response", exclude = c("pct_conf", "fwi_90th", "mean_evi"))
fv1

ggplot() +
 geom_line(data = fv1, aes(x = cmi_wy, y =.fitted)) +
   scale_fill_viridis_c(name = "Predicted\nMedian Burn Severity") +
   labs(x = "VPD (z-score)", y = "Lag (months)") +
   theme_minimal()
```