```{r}
library(ggplot2)
library(patchwork)
library(gratia)
# source
source(here("utils", "plot_functions.r"))
```


## Fuels plots

We plotted the linear predictor from the CMI model because it had slightly better fit for both median and extremes

```{r}

# For median burn severity (m_cmi_median)
# Get estimates for each ecoregion's pct_conf smooth
smooth_med <- smooth_estimates(m_median, 
  select = "s(pct_conf)", 
  n = 100, partial = TRUE, 
  unconditional = TRUE) |> add_confint()




# Plot with overlaid lines per ecoregion
conf_med <- ggplot() +
  geom_line(data = smooth_med, aes(x = pct_conf, y = .estimate, ), size = 1.5, lineend = "round") +
  geom_ribbon(data = smooth_med, aes(x = pct_conf, ymin = .lower_ci, ymax = .upper_ci), alpha = 0.2) +
  labs(x = "Percent Conifer Cover",  y = "Median Burn Severity") +
  theme_bw() 



# For extreme burn severity (m_cmi_ext)
smooth_ext <- smooth_estimates(m_ext, 
  select = "s(pct_conf)",
     n = 100, partial = TRUE, 
      unconditional = TRUE) |> add_confint()

conf_ext <- ggplot() +
  #geom_point(data = moisture_data, aes(x = pct_conf, y = pct_95, color = ecoregion_lvl), alpha = 0.1) +
  geom_line(data = smooth_ext, aes(x = pct_conf, y = .estimate), size = 1.5, lineend = "round") +
  geom_ribbon(data = smooth_ext, aes(x = pct_conf, ymin = .lower_ci, ymax = .upper_ci), alpha = 0.2) +
  labs(x = "Percent Conifer Cover",  y = "Extreme Burn Severity") +
  theme_bw() 
conf_ext


```

# Fire weather

```{r}

# For median burn severity (m_cmi_median)
# Get estimates for each ecoregion's pct_conf smooth
smooth_med <- smooth_estimates(m_median, 
  select = "s(dc_90th)", 
    n = 100, partial = TRUE, 
    unconditional = TRUE) |> add_confint()
smooth_ext<- smooth_estimates(m_ext, 
  select = "s(dc_90th)", 
  n = 100, 
  partial = TRUE,
  unconditional = TRUE)  |> add_confint()



# Plot with overlaid lines per ecoregion
fwi_med<- ggplot() +
  #geom_point(data = moisture_data, aes(x = mean_evi, y = median, color = ecoregion_lvl), alpha = 0.1) +
  geom_line(data = smooth_med, aes(x = dc_90th, y = .estimate), size = 1.5, lineend = "round") +
  geom_ribbon(data = smooth_med, aes(x = dc_90th, ymin = .lower_ci, ymax = .upper_ci), alpha = 0.2) +
  labs(x = "DC 90th Percentile", y = "Median Burn Severity") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) 

fwi_ext <- ggplot() +
  #geom_point(data = moisture_data, aes(x = mean_evi, y = median, color = ecoregion_lvl), alpha = 0.1) +
  geom_line(data = smooth_ext, aes(x = dc_90th, y = .estimate), size = 1.5, lineend = "round") +
  geom_ribbon(data = smooth_ext, aes(x = dc_90th, ymin = .lower_ci, ymax = .upper_ci), alpha = 0.2) +
  labs(x = "DC 90th Percentile",  y = "Extreme Burn Severity") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) 



```


```{r}
conf_med + conf_ext + fwi_med + fwi_ext + plot_layout(ncol = 2)  + plot_annotation(tag_levels = "A") 

# save
ggsave(here("figures", "smooth_predictor_plots.png"), width = 6, height = 6)
```


# Deviance explained by each group

```{r}
var_colours <- c("CMI" = "gray30", "VPD" = "orangered", "Drought Code" = "firebrick", "Conifer Cover" = "darkgreen")

I.ext <- extreme_hp$hierarchical.partitioning |> 
    as.data.frame() |> 
    rownames_to_column(var = "Variable") |> mutate(Variable = 
    case_when(
        Variable == "fire_weather" ~ "Drought Code", 
        Variable == "fuels" ~ "Conifer Cover",
        Variable == "intra" ~ "VPD",
        Variable == "inter" ~ "CMI",
        ))

I.median <- median_hp$hierarchical.partitioning |> 
    as.data.frame() |> 
    rownames_to_column(var = "Variable") |> mutate(Variable = 
    case_when(
        Variable == "fire_weather" ~ "Drought Code", 
        Variable == "fuels" ~ "Conifer Cover",
        Variable == "intra" ~ "VPD",
        Variable == "inter" ~ "CMI",
        ))

p.med <- ggplot(I.median, aes(x = Variable, y = `I.perc(%)`, fill = Variable)) +
  geom_col() +
  geom_text(aes(label = paste0(round(`I.perc(%)`, 1), "%")), vjust = -0.5, size = 3.5) +
  labs(title = "Median", y = "Deviance Explained (%)", x = "") +
  theme_bw() +
  coord_cartesian(ylim = c(0, 100)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        plot.title = element_text(hjust = 0.5, size = 12, face = "bold")) +
  scale_fill_manual(values = var_colours) +
  guides(fill = "none")

p.ext <- ggplot(I.ext, aes(x = Variable, y = `I.perc(%)`, fill = Variable)) +
  geom_col() +
  geom_text(aes(label = paste0(round(`I.perc(%)`, 1), "%")), vjust = -0.5, size = 3.5) +
  labs(title = "Extreme", y = NULL, x = "") +
  theme_bw() +
  coord_cartesian(ylim = c(0, 100)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        plot.title = element_text(hjust = 0.5, size = 12, face = "bold")) +
  scale_fill_manual(values = var_colours) +
  guides(fill = "none")



# combine plots using patchwork
 p.med + p.ext + plot_layout(ncol = 2) + 
  plot_annotation(tag_levels = "A")

# save
ggsave(here("figures", "deviance_explained_by_variable_group_v2.png"), width = 5, height = 4)


```

# H4: are there lagged effects?

get draws
```{r}

# vpd median
median_5_v <- one_d_slice_draws(m_median, smooth_id = 2, q = "5%", moisture_data = moisture_data)
median_25_v <- one_d_slice_draws(m_median, smooth_id = 2, q = "25%", moisture_data = moisture_data)
median_50_v <- one_d_slice_draws(m_median, smooth_id = 2, q = "50%", moisture_data = moisture_data)
median_75_v <- one_d_slice_draws(m_median, smooth_id = 2, q = "75%", moisture_data = moisture_data)

# cmi median
median_5_cmi <- one_d_slice_draws(m_median, smooth_id = 1, q = "5%", moisture_data = moisture_data)
median_25_cmi <- one_d_slice_draws(m_median, smooth_id = 1, q = "25%", moisture_data = moisture_data)
median_50_cmi <- one_d_slice_draws(m_median, smooth_id = 1, q = "50%", moisture_data = moisture_data)
median_75_cmi <- one_d_slice_draws(m_median, smooth_id = 1, q = "75%", moisture_data = moisture_data)

# vpd extreme
extreme_5_v <- one_d_slice_draws(m_ext, smooth_id = 2, q = "5%", moisture_data = moisture_data)
extreme_25_v <- one_d_slice_draws(m_ext, smooth_id = 2, q = "25%", moisture_data =  moisture_data)
extreme_50_v <- one_d_slice_draws(m_ext, smooth_id = 2  , q = "50%", moisture_data = moisture_data)
extreme_75_v <- one_d_slice_draws(m_ext, smooth_id = 2, q = "75%", moisture_data = moisture_data)

#cmi extreme
extreme_5_cmi <- one_d_slice_draws(m_ext, smooth_id = 1, q = "5%", moisture_data = moisture_data)
extreme_25_cmi <- one_d_slice_draws(m_ext, smooth_id = 1, q = "25%", moisture_data = moisture_data)
extreme_50_cmi <- one_d_slice_draws(m_ext, smooth_id = 1, q = "50%", moisture_data = moisture_data)
extreme_75_cmi <- one_d_slice_draws(m_ext, smooth_id = 1, q = "75%", moisture_data = moisture_data)

```

```{r}
# rbind all vpd median
median_vpd_all <- rbind(median_5_v$median,
                        median_25_v$median,
                        median_50_v$median,
                        median_75_v$median)
plot_colours <- c("5%" = "gray30", "25%" = "orangered", "50%" = "firebrick", "75%" = "darkgreen")
median_vpd_all$quartile <- factor(median_vpd_all$quartile, levels = c("5%", "25%", "50%", "75%"))

# Plot (unchanged, but legend will now be in this order)
vpd_m_q <- ggplot(median_vpd_all, aes(x = L, y = fit, color = quartile)) +
  geom_line(linewidth = 1) +
  labs(x = "Lag (days before fire)", y = "Partial Effect on Median Severity", color = "Quartile") +
  scale_x_reverse(breaks = seq(1,30,5), labels = paste0("-", seq(1,30,5))) +
  scale_color_manual(values = plot_colours) +
  theme_bw() +
  theme(
    panel.border = element_rect(fill = NA, colour = "black", size = 1.5),
    text = element_text(family = "Barlow", size = 14)
  )

# rbind all vpd extreme
extreme_vpd_all <- rbind(extreme_5_v$median,
                        extreme_25_v$median,
                        extreme_50_v$median,
                        extreme_75_v$median)
extreme_vpd_all$quartile <- factor(extreme_vpd_all$quartile, levels = c("5%", "25%", "50%", "75%"))

# Plot (unchanged, but legend will now be in this order)
vpd_e_q <- ggplot(extreme_vpd_all, aes(x = L, y = fit, color = quartile)) +
  geom_line(linewidth = 1) +
  labs(x = "Lag (days before fire)", y = "Partial Effect on Extreme Severity", color = "Quartile") +
  scale_x_reverse(breaks = seq(1,30,5), labels = paste0("-", seq(1,30,5))) +
  scale_color_manual(values = plot_colours) +
  theme_bw() +
  theme(
    panel.border = element_rect(fill = NA, colour = "black", size = 1.5),
    text = element_text(family = "Barlow", size = 14)
  )

vpd_m_q + vpd_e_q 
```


```{r}
# rbind all cmi median
median_cmi_all <- rbind(median_5_cmi$median,
                        median_25_cmi$median,
                        median_50_cmi$median,
                        median_75_cmi$median)
median_cmi_all$quartile <- factor(median_cmi_all$quartile, levels = c("5%", "25%", "50%", "75%"))

# Plot (unchanged, but legend will now be in this order)
cmi_m_q <- ggplot(median_cmi_all, aes(x = LL, y = fit, color = quartile)) +
  geom_line(linewidth = 1) +
  labs(x = "Lag (years before fire)", y = "Partial Effect on Median Severity", color = "Quartile") +
  scale_x_reverse(breaks = 1:5, labels = paste0("-", 1:5)) +
  scale_color_manual(values = plot_colours) +
  theme_bw() +
  theme(
    panel.border = element_rect(fill = NA, colour = "black", size = 1.5),
    text = element_text(family = "Barlow", size = 14)
  )
# rbind all cmi extreme
extreme_cmi_all <- rbind(extreme_5_cmi$median,
                        extreme_25_cmi$median,
                        extreme_50_cmi$median,
                        extreme_75_cmi$median)
extreme_cmi_all$quartile <- factor(extreme_cmi_all$quartile, levels = c("5%", "25%", "50%", "75%"))

# Plot (unchanged, but legend will now be in this order)
cmi_e_q <- ggplot(extreme_cmi_all, aes(x = LL, y = fit, color = quartile)) +
  geom_line(linewidth = 1) +
  labs(x = "Lag (years before fire)", y = "Partial Effect on Extreme Severity", color = "Quartile") +
  scale_x_reverse(breaks = 1:5, labels = paste0("-", 1:5)) +  
  scale_color_manual(values = plot_colours) +
  theme_bw() +
  theme(
    panel.border = element_rect(fill = NA, colour = "black", size = 1.5),
    text = element_text(family = "Barlow", size = 14)
  )
cmi_m_q + cmi_e_q
```
Analyzed lagged effects through visualization of the 2D smooths from the distributed lag smooths. Significant areas circled in black . If no significant areas, there were no differences across lags. Suggests that lagged effects are not as importance as the overall effects of moisture condditions on fire severity.


# 2D Smooths
 

## Median
```{r}


estimates_med <- get_lag_estimates(m_median, lag_var = "L", smooth_var = "vpd")
plot_vpd_med <- plot_lag_smooth(m_median, estimates_med, "vpd", "L")
plot_vpd_med

estimates_med_c <- get_lag_estimates(m_median, lag_var = "LL", smooth_var = "cmi_wy")
plot_c_med <- plot_lag_smooth(m_median, estimates_med_c, "cmi_wy", "LL")
plot_c_med 



```

## extreme
```{r}
estimates_ext <- get_lag_estimates(m_ext, lag_var = "L", smooth_var = "vpd")
plot_vpd_ext <- plot_lag_smooth(m_ext, estimates_ext, "vpd", "L")
plot_vpd_ext

estimates_ext_c <- get_lag_estimates(m_ext, lag_var = "LL", smooth_var = "cmi_wy")
plot_c_ext <- plot_lag_smooth(m_ext, estimates_ext_c, "cmi_wy", "LL")
plot_c_ext 


```


patchwork them together median plot with VPD on left, CMI on right with 2 Col, one row 2d smooths and one for quartile
```{r}

# median
p_m <- plot_vpd_med + plot_c_med + plot_layout(ncol = 2) 
q_m <- vpd_m_q + cmi_m_q + plot_layout(ncol = 2) + plot_layout(widths = c(2, 2))+ plot_layout(guides = 'collect') & theme(legend.position = 'bottom')



p1 <- (p_m / q_m) + plot_layout(ncol = 1) + plot_annotation(tag_levels = "A")

p1

#save to figures
ggsave(here("figures", "lagged_effects_median_v5.png"), width = 8, height = 8, dpi = 300)

# extreme
p_e <- plot_vpd_ext + plot_c_ext + plot_layout(ncol = 2) 
q_e <- vpd_e_q + cmi_e_q + plot_layout(ncol = 2) + plot_layout(widths = c(2, 2))+ plot_layout(guides = 'collect') & theme(legend.position = 'bottom')
p2 <- (p_e / q_e) + plot_layout(ncol = 1) + plot_annotation(tag_levels = "A")
p2
#save to figures
p3 <- (plot_vpd_med + plot_c_med + plot_vpd_ext + plot_c_ext ) + plot_layout(ncol = 2) + plot_annotation(tag_levels = "A")
ggsave(here("figures", "lagged_effects_median_extreme.png"), width = 8, height = 8, dpi = 300)


p4 <- cmi_m_q + cmi_e_q + plot_layout(ncol = 2) +  plot_layout(guides = 'collect') 
ggsave(here("figures", "cmi_lagged_effects_median_extreme.png"), width = 8, height = 4, dpi = 300)

```

# steep slops


```{r}
# median vs extreme 
get_steepest(estimates_med)    
get_steepest(estimates_med_c)

get_steepest(estimates_ext)
get_steepest(estimates_ext_c)

summary(get_steepest(estimates_med)$sig)
```

